### 2025-01-31 products/manuscript/Manuscript.qmd
Part 1

---
title: "The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer:A SEER-Based Cohort Study"
author: "Mohammed Zuber"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

* Mohammed Zuber$^{1,}$ (ORCID: 0000-0000-1234-5678)

**Author affiliations**

1. College of Pharmacy, University of Georgia, Athens, GA, USA.

$\land$ Corresponding author: mohammedzuber099@gmail.com

$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
This study will utilize the Surveillance, Epidemiology, and End Results (SEER) database, one of the most comprehensive cancer incidence and survival registries, covering over 48% of the United States. SEER collects key demographic data, cancer characteristics, treatment details, and survival outcomes. Managed by the U.S. National Cancer Institute, the database began recording information on lung, bone, liver, and brain metastases in 2010. Data for this study will be accessed through SEER*Stat 8.4.4 software, using the SEER Research Data (2000–2019) from 17 registries.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

_Question_ What are the clinicopathological features and survival outcomes of patients with different metastatic sites in stage IV endometrial cancer?

_Hypothesis_ Patients with stage IV endometrial cancer exhibit distinct clinicopathological characteristics and survival outcomes based on the site of metastasis. Specifically, metastases to different organ sites (e.g., lung, liver, bone, and brain) are associated with varying prognostic implications, with some metastatic patterns demonstrating significantly poorer survival compared to others. By analyzing data from the SEER database, we hypothesize that certain metastatic sites will be linked to more aggressive tumor behavior and worse overall survival, highlighting potential differences in disease progression and treatment response.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
This study examines survival outcomes in stage IV endometrial cancer using SEER data. The primary outcomes include survival months, cause-specific death classification, and vital status at study cutoff. Key predictors include age, race, income, year of diagnosis, and presence of differnt metastasis status.

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._
 The analysis aims to identify survival differences based on metastatic patterns, demographic disparities, socioeconomic influences, and temporal trends in prognosis. Kaplan-Meier survival analysis will used to compare the survival among the study cohorts. The incidence of metastasis will be calculated in stage IV endometrial cancer. A Cox regression model will be performed to compute the adjusted Hazard Ratios (aHR) and 95% confidence intervals (CI).

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-02-09 products/manuscript/Manuscript.qmd
Added more description based on Dr. @andreashandel comments. #1

@@
 _Provide enough background on your topic that others can understand the why and how of your analysis_ 
+Endometrial cancer, a neoplasm originating from the endometrial lining, is the most common gynecological cancer in the United States.[@Monk2022] According to American Cancer Society 2024 statistics the incidence of uterine cancer is 67, 880 with 13,250 deaths.
+Due to its early symptomatic presentation, 70% of cases are diagnosed at stage I, where the five-year survival rate exceeds 90%.
+However, around 13% of patients are diagnosed at advanced stages (III/IV).
+
+A retrospective study found that the prevalence of stage IVB endometrial cancer increased to 5.67% between 2010 and 2015.
+Unlike stage I, which has a favorable prognosis, stage IV is associated with poor outcomes, with a five-year survival rate ranging from 17% to 30%.
+The most frequent metastatic sites include the peritoneum, lungs, pelvic and para-aortic lymph nodes, and local pelvic recurrences.
+Among these, lung metastasis is one of the most common distant metastases, occurring in 1.5% of patients.
+The hematogenous spread is the primary mechanism of lung metastasis.
+Depending on clinical features, disease stage, prior surgical history, and pathological factors, 11-13% of endometrial cancer patients experience recurrence within the first two years after initial treatment.
+Symptoms of lung metastasis typically include coughing up blood, shortness of breath, chest pain, persistent cough, pleural effusion, reduced appetite, and weight loss.
+
+Compared to other gynecological cancers such as cervical and ovarian cancer, endometrial carcinoma exhibits the highest rate of pulmonary metastasis, affecting up to 20-25% of patients who experience recurrence.
+According to Mao et al., 2020, the five-year cancer-specific survival (CSS) rate for liver and brain metastases was 21%, while bone metastases had a survival rate of 20%.
+In patients with lung metastases, the median overall survival was 11 months, with a five-year CSS rate of 19%.
+However, this data specifically pertains to stage IV lung cancer patients.
+
+Patients with advanced endometrial cancer face poor survival outcomes, with metastatic disease being the leading cause of mortality.
+Therefore, this study aimed to utilize data from the Surveillance, Epidemiology, and End Results (SEER) database to analyze survival outcomes in patients diagnosed with stage IV endometrial cancer.
+The primary objective was to determine whether the presence of lung/brain/bone/liver metastases in adults with stage IV endometrial cancer is associated with worse survival outcomes compared to patients without any metastasis.
+
+Please note the references are not yet added and will be added soon.
 
@@
 Data for this study will be accessed through SEER*Stat 8.4.4 software, using the SEER Research Data (2000–2019) from 17 registries.
+
+The extracted variables include age, race, income, American Joint Committee on Cancer (AJCC) stage, year of cancer diagnosis, lung metastasis (yes/no), brain metastasis (yes/no), liver metastasis(yes/no), brain metastasis (yes/no), cause of death, survival months, vital status (dead/alive), rural/urban classification.
 
@@
 _As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+The data was extracted from SEER*Stat 8.4.4 software, using the SEER Research Data (2000–2019) from 17 registries.
+The following inclusion criteria was used to extract the data (1) primary site, C54.1–9 and C55.9 based on International Classification of Tumor Diseases for Oncology (ICD-O) codes; (2) year of diagnosis 2010-2015; (3) AJCC stage IV; (4) age more than 18 years.
+
+The SEER database provided data on both demographic and clinical characteristics of the patients.
+The demographic variables included age at diagnosis, race, median household income (inflation-adjusted to 2019), year of diagnosis, and Derived AJCC Stage Group (7th edition, 2010-2015).
+The clinical variables comprised lung, brain, bone, and liver metastasis status, survival duration (in months), SEER cause-specific death classification, and vital status recode (used for study cutoff determination).
+
+Age will be categorized into four groups: 18-39, 40-59, 60-79, and ≥80 years.
+Race is classified into Asian or Pacific Islander, American Indian or Alaska Native, Black, and White.
+The income variable will be divided into six brackets: < $35,000, $35,000 - $44,999, $45,000 - $54,999, $55,000 - $64,999, $65,000 - $74,999, and $75,000+.
+The SEER cause-specific death classification was preferred over all-cause mortality to ensure that only deaths specifically attributed to endometrial cancer will be included in the analysis.
 
@@
 All materials, including files that contain code, should be commented well so everyone can follow along._
+
+Endometrial cancer cases will be extracted from the SEER database.
+Patients with an unknown lung/brain/bone/liver metastasis status will be excluded.
+Individuals under 18 years of age will be removed from the dataset.
+Additionally, patients whose cause of death is unrelated to endometrial cancer or whose cause of death is missing or unknown will not be included in the analysis.
+Cases diagnosed at AJCC stages I to III will be excluded from the study.
+The final cohort will be categorized into 5 groups:  lung metastasis, brain metastasis, bone metastasis, liver metastasis and no metastasis.
+
+The data cleaning process has not yet started. The R code will be provided as the analysis completed.
 

### 2025-02-09 products/manuscript/Manuscript.qmd
Added more description based on Dr. @adreashandel comments. V2 #1

@@
 _Provide enough background on your topic that others can understand the why and how of your analysis_ 
-Endometrial cancer, a neoplasm originating from the endometrial lining, is the most common gynecological cancer in the United States.[@Monk2022] According to American Cancer Society 2024 statistics the incidence of uterine cancer is 67, 880 with 13,250 deaths.
+Endometrial cancer, a neoplasm originating from the endometrial lining, is the most common gynecological cancer in the United States.
+According to American Cancer Society 2024 statistics the incidence of uterine cancer is 67, 880 with 13,250 deaths.
 Due to its early symptomatic presentation, 70% of cases are diagnosed at stage I, where the five-year survival rate exceeds 90%.

### 2025-02-21 code/eda-code/edamz.qmd
part 2 project

---
title: "Exploratory analysis script"
date: "2024-02-20"
output: html_document
---

The purpose of this script is to conduct an exploratory analysis of the given dataset. This involves summarizing key variables, identifying patterns, and visualizing data distributions. The analysis will help in understanding the dataset’s structure, detecting anomalies, and forming preliminary insights for further modeling or hypothesis testing.

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

Generate a summary of the dataset using the skimr package and print it for review.  The summary provides key statistics such as missing values, variable types, and distributions. The output is saved as an RDS file in the "results/edatables/" directory.  The prefix "eda" in the filename helps distinguish exploratory analysis files from final analysis outputs.  Storing the summary table allows for future reference without the need to rerun the analysis.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "edatables", "edasummarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

I am exploring the variable age first with the table

```{r}
# Summary of Age recode variable (categorical)
p1 <- mydata %>%
  group_by(`Age recode with <1 year olds`) %>%
  summarise(Count = n())

# Print the summary table
print(p1)

# Save the summary table
summarytable_file = here("results", "edatables", "edaagesummary.rds")
saveRDS(p1, file = summarytable_file)

```
Now lets look at the bar plot for age distribution

```{r}
# Bar plot for Age recode distribution (categorical)
p2 <- mydata %>%
  ggplot(aes(x = `Age recode with <1 year olds`)) + 
  geom_bar(fill = "skyblue") + 
  labs(title = "Age Distribution (Categorical)", x = "Age Group", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for readability

# Display the plot
plot(p2)

# Save the plot
figure_file = here("results", "edafigures", "edaagedistribution.png")
ggsave(filename = figure_file, plot = p2)

```
Now, exploring the Race recode (W, B, AI, API) variable

```{r}
# Summary of Race recode variable (categorical)
# - Groups the dataset by the "Race recode (W, B, AI, API)" variable.
# - Counts the number of observations in each race category.
p3 <- mydata %>%
  group_by(`Race recode (W, B, AI, API)`) %>%
  summarise(Count = n())

# Print the summary table
print(p3)

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edaracesummary.rds")
saveRDS(p3, file = summarytable_file)
```
```{r}
# Bar plot for Race recode distribution (categorical)
# - Visualizes the distribution of race categories.
# - Uses "skyblue" fill for better readability.
p4 <- mydata %>%
  ggplot(aes(x = `Race recode (W, B, AI, API)`)) + 
  geom_bar(fill = "skyblue") + 
  labs(title = "Race Distribution", x = "Race Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for better readability

# Display the plot
plot(p4)

# Save the plot
# - Stored in "results/edafigures/" with the prefix "eda" to indicate exploratory analysis.
figure_file = here("results", "edafigures", "edaracedistribution.png")
ggsave(filename = figure_file, plot = p4)

```
Now exploring Year of Diagnosis 
```{r}
# Summary of Year of Diagnosis variable (categorical)
# - Groups the dataset by "Year of diagnosis" and counts occurrences.
p5 <- mydata %>%
  group_by(`Year of diagnosis`) %>%
  summarise(Count = n())

# Print the summary table
print(p5)

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edayearsummary.rds")
saveRDS(p5, file = summarytable_file)

```
```{r}
# Bar plot for Year of Diagnosis distribution (categorical)
# - Visualizes the distribution of diagnosis years.
# - Uses "skyblue" fill for better readability.
p6 <- mydata %>%
  ggplot(aes(x = factor(`Year of diagnosis`))) +  # Treat as categorical
  geom_bar(fill = "skyblue") + 
  labs(title = "Year of Diagnosis Distribution", x = "Year", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotates x-axis labels for better readability

# Display the plot
plot(p6)

# Save the plot
# - Stored in "results/edafigures/" with the prefix "eda" to indicate exploratory analysis.
figure_file = here("results", "edafigures", "edayear.png")
ggsave(filename = figure_file, plot = p6)

```
Now, lets explore SEER Combined Mets at DX (Bone, Brain, Liver, Lung, and Distant LN) using p7 for the summary table and p8 for the bar plot

```{r}
# Summary of SEER Combined Mets at DX variables (categorical)
# - Groups data by metastasis categories and counts occurrences.
p7 <- mydata %>%
  summarise(
    Bone_Mets = sum(`SEER Combined Mets at DX-bone (2010+)` == "Yes", na.rm = TRUE),
    Brain_Mets = sum(`SEER Combined Mets at DX-brain (2010+)` == "Yes", na.rm = TRUE),
    Liver_Mets = sum(`SEER Combined Mets at DX-liver (2010+)` == "Yes", na.rm = TRUE),
    Lung_Mets = sum(`SEER Combined Mets at DX-lung (2010+)` == "Yes", na.rm = TRUE),
      ) %>%
  pivot_longer(cols = everything(), names_to = "Metastasis_Site", values_to = "Count")

# Print the summary table
print(p7)

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edametsatsummary.rds")
saveRDS(p7, file = summarytable_file)

```

```{r}
# Bar plot for SEER Combined Mets at DX distribution
# - Visualizes the number of occurrences for each metastasis type.
# - Uses "skyblue" fill for better readability.
p8 <- p7 %>%
  ggplot(aes(x = Metastasis_Site, y = Count, fill = Metastasis_Site)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Metastasis at Diagnosis Distribution", x = "Metastasis Site", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  guides(fill = "none")  # Removes legend since it's redundant

# Display the plot
plot(p8)

# Save the plot
# - Stored in "results/edafigures/" with the prefix "eda" to indicate exploratory analysis.
figure_file = here("results", "edafigures", "edametsatdxdistribution.png")
ggsave(filename = figure_file, plot = p8)

```
Now, lets explore SEER cause-specific death classification

```{r}
# Summary of SEER Cause-Specific Death Classification variable (categorical)
# - Counts the number of occurrences for each cause-specific death classification category.
# - Also calculates the percentage for each category.
p9 <- mydata %>%
  group_by(`SEER cause-specific death classification`) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(Percentage = round((Count / sum(Count)) * 100, 1))

# Print the summary table
print(p9)

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edacausedeathsummary.rds")
saveRDS(p9, file = summarytable_file)

```

```{r}
# Bar plot for SEER Cause-Specific Death Classification distribution
# - Visualizes the distribution of cause-specific death classification.
p10 <- p9 %>%
  ggplot(aes(x = `SEER cause-specific death classification`, y = Count, fill = `SEER cause-specific death classification`)) + 
  geom_bar(stat = "identity") + 
  labs(title = "SEER Cause-Specific Death Classification", x = "Cause-Specific Death Category", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  guides(fill = "none")  # Removes legend since x-axis labels already describe the categories

# Display the plot
plot(p10)

# Save the plot
# - Stored in "results/edafigures/" with the prefix "eda" to indicate exploratory analysis.
figure_file = here("results", "edafigures", "edacausedeathdistribution.png")
ggsave(filename = figure_file, plot = p10)

```
Now, lets explore Survival Months

```{r}
# Summary of Survival Months variable (continuous)
# - Computes descriptive statistics for survival months.
p11 <- mydata %>%
  summarise(
    Min_Survival = min(`Survival months`, na.rm = TRUE),
    Max_Survival = max(`Survival months`, na.rm = TRUE),
    Mean_Survival = mean(`Survival months`, na.rm = TRUE),
    Median_Survival = median(`Survival months`, na.rm = TRUE),
    SD_Survival = sd(`Survival months`, na.rm = TRUE),
    Count = n()
  )

# Print the summary table
print(p11)

# Save the summary table
# - Stored in "results/edatables/" with the prefix "eda" to indicate exploratory analysis.
summarytable_file = here("results", "edatables", "edasurvivalmonthssummary.rds")
saveRDS(p11, file = summarytable_file)

```
```{r}
# Histogram for Survival Months distribution
# - Visualizes the distribution of survival months.
p12 <- mydata %>%
  ggplot(aes(x = `Survival months`)) + 
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 
  labs(title = "Survival Months Distribution", x = "Survival Months", y = "Count") +
  theme_minimal()

# Display the plot
plot(p12)

# Save the plot
# - Stored in "results/edafigures/" with the prefix "eda" to indicate exploratory analysis.
figure_file = here("results", "edafigures", "edasurvivalmonthsdistribution.png")
ggsave(filename = figure_file, plot = p12)

```

### 2025-02-21 code/processing-code/processingcodemz.qmd
part 2 project

---
title: "Cleaning script"
author: "Mohammed Zuber"
format: html
date: "2-20-2025""
editor: visual
---

# Processing script

This document contains all the code and a detailed description of the data cleaning process involved in this project. It provides explanations and comments for each step to facilitate understanding and reproducibility in the future.

This Quarto file integrates both code and text, allowing for a structured approach to data cleaning, preprocessing, and transformation.

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

# data loading
I am loading the raw data for further cleaning and processing 
```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","rawdata.csv")
rawdata <- readr::read_csv(data_location)
```
# Check Data

## First, load the codebook

```{r}
codebook_location <- here::here("data", "raw-data", "codebook.csv")
codebook <- readr::read_csv(codebook_location)
print(codebook)
```

## Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)  # Overview of data structure
summary(rawdata)  # Summary statistics
head(rawdata)  # First few rows of the dataset
skimr::skim(rawdata)  # Detailed summary including missing values
```

# Data Cleaning

By inspecting the data as done above, we identify some issues that need to be addressed.

Some variables were incorrectly read as numeric when they should be categorical (character).
Convert the following:
Year of diagnosis → Convert to character (since it represents a categorical year label rather than a numeric value).
RX Summ--Surg Prim Site (1998+) → Convert to character (since it represents surgery codes rather than numerical values).
Patient ID → Convert to character (since it's an identifier and should not be used in numerical calculations).
Survival months → Convert to numeric (since it was loaded as a character string and should be a numerical value for analysis).

## Fixing Data Types

```{r}
d1 <- rawdata %>% 
  dplyr::mutate(
    `Year of diagnosis` = as.character(`Year of diagnosis`),  # Convert to character
    `RX Summ--Surg Prim Site (1998+)` = as.character(`RX Summ--Surg Prim Site (1998+)`),  # Convert to character
    `Patient ID` = as.character(`Patient ID`),  # Convert to character
    `Survival months` = as.numeric(`Survival months`)  # Convert to numeric
  )

skimr::skim(d1)  # Check changes
```
## Converting Categorical Variables to Factor

To ensure categorical variables are properly treated, we convert several key variables to factor type. This is important for:
- Efficient storage and manipulation of categorical data.
- Enabling statistical models and visualizations to properly recognize categories.
- Improving the interpretability of categorical variables in analysis.

We apply this conversion to the following variables:
- **Sex** and **Race-related variables** to ensure demographic information is handled correctly.
- **Cancer staging and metastasis indicators** to ensure proper classification.
- **Cause of death and survival indicators** to maintain categorical consistency.
```{r}
d2 <- d1 %>% dplyr::mutate(
    Sex = as.factor(Sex),
    `Race recode (W, B, AI, API)` = as.factor(`Race recode (W, B, AI, API)`),
    `Origin recode NHIA (Hispanic, Non-Hisp)` = as.factor(`Origin recode NHIA (Hispanic, Non-Hisp)`),
    `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)` = as.factor(`Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`),
    `Derived AJCC Stage Group, 7th ed (2010-2015)` = as.factor(`Derived AJCC Stage Group, 7th ed (2010-2015)`),
    `SEER Combined Mets at DX-bone (2010+)` = as.factor(`SEER Combined Mets at DX-bone (2010+)`),
    `SEER Combined Mets at DX-brain (2010+)` = as.factor(`SEER Combined Mets at DX-brain (2010+)`),
    `SEER Combined Mets at DX-liver (2010+)` = as.factor(`SEER Combined Mets at DX-liver (2010+)`),
    `SEER Combined Mets at DX-lung (2010+)` = as.factor(`SEER Combined Mets at DX-lung (2010+)`),
    `Mets at DX-Distant LN (2016+)` = as.factor(`Mets at DX-Distant LN (2016+)`),
    `COD to site recode` = as.factor(`COD to site recode`),
    `SEER cause-specific death classification` = as.factor(`SEER cause-specific death classification`),
    `SEER other cause of death classification` = as.factor(`SEER other cause of death classification`),
    `Survival months flag` = as.factor(`Survival months flag`),
    `COD to site rec KM` = as.factor(`COD to site rec KM`),
    `Vital status recode (study cutoff used)` = as.factor(`Vital status recode (study cutoff used)`)
)

skimr::skim(d2)
```
## Removing Patients in Specific Age Categories

To focus our analysis on the adult patient population, we remove patients in the following age categories:
- **01-04 years**
- **15-19 years**

```{r}
d3 <- d2 %>% dplyr::filter(!`Age recode with <1 year olds` %in% c("01-04 years", "15-19 years"))
skimr::skim(d3)
```
## Including Patients with Stage IVA and IVB

To refine our dataset further, we include only patients who are categorized as **Stage IVA or IVB** based on the `Derived AJCC Stage Group, 7th ed (2010-2015)` column.

```{r}
d4 <- d3 %>% dplyr::filter(`Derived AJCC Stage Group, 7th ed (2010-2015)` %in% c("IVA", "IVB"))
skimr::skim(d4)
```
## Excluding Patients with Specific SEER Cause-Specific Death Classification

To ensure data quality, we exclude patients with `SEER cause-specific death classification` categorized as:
- **Dead (missing/unknown COD)**
- **N/A not seq 0-59**

```{r}
d5 <- d4 %>% dplyr::filter(!`SEER cause-specific death classification` %in% c("Dead (missing/unknown COD)", "N/A not seq 0-59"))
skimr::skim(d5)
```

## Excluding Patients with Unknown or Missing Survival Months

To ensure accurate survival analysis, we exclude patients who have **unknown or missing** survival months.

```{r}
d6 <- d5 %>% dplyr::filter(!is.na(`Survival months`))
skimr::skim(d6)
```
## All done, data is clean now.

Let's assign the final cleaned dataset to a variable. This makes it easier to add further cleaning steps above if needed.

```{r}
processeddata <- d6  # Assign cleaned dataset
```

# Save Data

Finally, we save the clean data as an RDS file. Saving as RDS or RDA/RData preserves coding like factors, characters, and numeric types, whereas CSV does not retain such metadata. However, CSV is useful for sharing as it is plain text. Consider documenting variable meanings if sharing as CSV.

For reference, see suggestions on storing processed data: [Saving Data into R Data Format](http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata)

```{r}
save_data_location <- here::here("data", "processed-data", "processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

# Notes

Removing individuals with "faulty" or missing data is one approach. However, depending on the question and analysis approach, different cleaning methods might be better (e.g., keeping individuals with some missing information).

### 2025-02-21 products/manuscript/Manuscript.qmd
part 2 project

--- 2025-01-31 products/manuscript/Manuscript.qmd
@@
 csl: ../../assets/american-journal-of-epidemiology.csl
+editor: 
+  markdown: 
+    wrap: sentence
 ---
 
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. 
+The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
+Adjust as needed.
 You don't need to have exactly these sections, but the content covering those sections should be addressed.
@@
 
-* Mohammed Zuber$^{1,}$ (ORCID: 0000-0000-1234-5678)
+-   Mohammed Zuber$^{1,}$ (ORCID: 0000-0000-1234-5678)
 
@@
 
-$\land$ Corresponding author: mohammedzuber099@gmail.com
+$\land$ Corresponding author: mohammedzuber099\@gmail.com
 
-$\dagger$ Disclaimer: The opinions expressed in this article are the
-author's own and don't reflect their employer.
+$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
 
@@
 # Summary/Abstract
-_Write a summary of your project._
+
+*Write a summary of your project.*
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+Endometrial cancer, a neoplasm originating from the endometrial lining, is the most common gynecological cancer in the United States.[@Monk2022; @MORICE20161094] According to American Cancer Society 2024 statistics the incidence of uterine cancer is 67, 880 with 13,250 deaths.
+Due to its early symptomatic presentation, 70% of cases are diagnosed at stage I, where the five-year survival rate exceeds 90%.
+[@MORICE20161094; @Galaal14] However, around 13% of patients are diagnosed at advanced stages (III/IV).[@Galaal14]
+
+A retrospective study found that the prevalence of stage IVB endometrial cancer increased to 5.67% between 2010 and 2015.[@Guo2020] Unlike stage I, which has a favorable prognosis, stage IV is associated with poor outcomes, with a five-year survival rate ranging from 17% to 30%.[@Carvalho2020] The most frequent metastatic sites include the peritoneum, lungs, pelvic and para-aortic lymph nodes, and local pelvic recurrences.[@DeNardo2023] Among these, lung metastasis is one of the most common distant metastases, occurring in 1.5% of patients.[@Mao2020] The hematogenous spread is the primary mechanism of lung metastasis.[@Ballon1979] Depending on clinical features, disease stage, prior surgical history, and pathological factors, 11-13% of endometrial cancer patients experience recurrence within the first two years after initial treatment.
+Symptoms of lung metastasis typically include coughing up blood, shortness of breath, chest pain, persistent cough, pleural effusion, reduced appetite, and weight loss.[@Jamil2025]
+
+Compared to other gynecological cancers such as cervical and ovarian cancer, endometrial carcinoma exhibits the highest rate of pulmonary metastasis, affecting up to 20-25% of patients who experience recurrence.
+According to Mao et al., 2020, the five-year cancer-specific survival (CSS) rate for liver and brain metastases was 21%, while bone metastases had a survival rate of 20%.
+In patients with lung metastases, the median overall survival was 11 months, with a five-year CSS rate of 19%.
+However, this data specifically pertains to stage IV lung cancer patients.[@Mao2020] Patients with advanced endometrial cancer face poor survival outcomes, with metastatic disease being the leading cause of mortality.
+[@Amant2005] Therefore, this study aimed to utilize data from the Surveillance, Epidemiology, and End Results (SEER) database to analyze survival outcomes in patients diagnosed with stage IV endometrial cancer.
+The primary objective was to determine whether the presence of lung/brain/bone/liver metastases in adults with stage IV endometrial cancer is associated with worse survival outcomes compared to patients without any metastasis.
 
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+
 This study will utilize the Surveillance, Epidemiology, and End Results (SEER) database, one of the most comprehensive cancer incidence and survival registries, covering over 48% of the United States.
@@
 National Cancer Institute, the database began recording information on lung, bone, liver, and brain metastases in 2010.
-Data for this study will be accessed through SEER*Stat 8.4.4 software, using the SEER Research Data (2000–2019) from 17 registries.
+Data for this study will be accessed through SEER\*Stat 8.4.4 software, using the SEER Research Data (2000–2019) from 17 registries.
+
+The extracted variables include age, race, income, American Joint Committee on Cancer (AJCC) stage, year of cancer diagnosis, lung metastasis (yes/no), brain metastasis (yes/no), liver metastasis (yes/no), brain metastasis (yes/no), cause of death, survival months, vital status (dead/alive), rural/urban classification.
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
 
-_Question_ What are the clinicopathological features and survival outcomes of patients with different metastatic sites in stage IV endometrial cancer?
+*Question* What are the clinicopathological features and survival outcomes of patients with different metastatic sites in stage IV endometrial cancer?
 
-_Hypothesis_ Patients with stage IV endometrial cancer exhibit distinct clinicopathological characteristics and survival outcomes based on the site of metastasis.
+*Hypothesis* Patients with stage IV endometrial cancer exhibit distinct clinicopathological characteristics and survival outcomes based on the site of metastasis.
 Specifically, metastases to different organ sites (e.g., lung, liver, bone, and brain) are associated with varying prognostic implications, with some metastatic patterns demonstrating significantly poorer survival compared to others.
 By analyzing data from the SEER database, we hypothesize that certain metastatic sites will be linked to more aggressive tumor behavior and worse overall survival, highlighting potential differences in disease progression and treatment response.
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
 This study examines survival outcomes in stage IV endometrial cancer using SEER data.
@@
 
-## Schematic of workflow
+## Data aquisition
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+The data was extracted from SEER\*Stat 8.4.4 software, using the SEER Research Data (2000–2019) from 17 registries.
+The following inclusion criteria was used to extract the data (1) primary site, C54.1–9 and C55.9 based on International Classification of Tumor Diseases for Oncology (ICD-O) codes; (2) year of diagnosis 2010-2015; (3) AJCC stage IV; (4) age more than 18 years.
 
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
+The SEER database provided data on both demographic and clinical characteristics of the patients.
+The demographic variables included age at diagnosis, race, median household income (inflation-adjusted to 2019), year of diagnosis, and Derived AJCC Stage Group (7th edition, 2010-2015).
+The clinical variables comprised lung, brain, bone, and liver metastasis status, survival duration (in months), SEER cause-specific death classification, and vital status recode (used for study cutoff determination).
 
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+Age will be categorized into four groups: 18-39, 40-59, 60-79, and ≥80 years.
+Race is classified into Asian or Pacific Islander, American Indian or Alaska Native, Black, and White.
+The income variable will be divided into six brackets: \< \$35,000, \$35,000 - \$44,999, \$45,000 - \$54,999, \$55,000 - \$64,999, \$65,000 - \$74,999, and \$75,000+.
+The SEER cause-specific death classification was preferred over all-cause mortality to ensure that only deaths specifically attributed to endometrial cancer will be included in the analysis.
 
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+
+Endometrial cancer cases were extracted from the SEER database.
+Individuals under 18 years of age were removed from the dataset.
+Additionally, patients whose cause of death is unrelated to endometrial cancer or whose cause of death is missing or unknown were not included in the analysis.
+Cases diagnosed at AJCC stages I to III was excluded from the study.
+The final cohort was categorized into 4 groups: lung metastasis, brain metastasis, bone metastasis, and liver metastasis.The SEER dataset underwent a structured cleaning process to ensure data integrity and suitability for analysis.
+First, data types were standardized by converting categorical variables such as Year of Diagnosis, RX Summ--Surg Prim Site (1998+), and Patient ID to character format, while Survival Months was converted to numeric.
+Key categorical variables, including Sex, Race, Metastasis Indicators, Survival Flags, and Cause-Specific Death Classification, were transformed into factors to maintain consistency.
+Patients within the age categories "01-04 years" and "15-19 years" were excluded, and only those classified as Stage IVA or IVB in the Derived AJCC Stage Group, 7th ed (2010-2015) were retained.
+Additionally, patients with "Dead (missing/unknown COD)" or "N/A not seq 0-59" in the SEER cause-specific death classification were removed.
+To ensure completeness, individuals with missing or unknown survival months were excluded.
+The final cleaned dataset was saved for further analysis, ensuring consistency and reproducibility.The code and raw data to reproduce the data cleaning step is available in https://github.com/mohammedzuber099/ZUBER-MADA-project.
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+
 The analysis aims to identify survival differences based on metastatic patterns, demographic disparities, socioeconomic influences, and temporal trends in prognosis.
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
 @tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
 
@@
 #| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
+resulttable=readRDS(here("results","edatables","edasummarytable.rds"))
 knitr::kable(resulttable)
+```
+
+@tbl-summarytablesurvival gives the overview of the survival.
+The survival months in our dataset range from 0 to 119 months, with a mean survival of 23.1 months and a median survival of 12 months, indicating a right-skewed distribution.
+The standard deviation (SD) is 27.67 months, reflecting high variability in patient survival.
+A total of 5,839 patients were included in the analysis.
+This distribution suggests that while some patients survive long-term, the majority have relatively short survival durations.
+
+```{r}
+#| label: tbl-summarytablesurvival
+#| tbl-cap: "Summary of Survival."
+#| echo: FALSE
+resulttable=readRDS(here("results","edatables","edasurvivalmonthssummary.rds"))
+knitr::kable(resulttable)
+```
+
+@fig-survival shows the distribution of survival months.
+This histogram illustrates the distribution of survival months among patients in the dataset.
+The x-axis represents survival duration in months, while the y-axis shows the count of patients within each survival interval.
+The distribution is right-skewed, indicating that most patients had shorter survival times, with a gradual decline in frequency as survival duration increases.
+A substantial proportion of patients survived less than 12 months, while fewer cases exhibited long-term survival beyond 60 months.
+The presence of some patients with survival exceeding 100 months highlights potential long-term survivors.
+
+```{r}
+#| label: fig-survival 
+#| fig-cap: "Survival months."
+#| echo: FALSE
+knitr::include_graphics(here("results","edafigures","edasurvivalmonthsdistribution.png"))
+```
+
+@tbl-summaryage and @fig-age shows age distribution.The dataset includes patients across a wide age range, with the majority falling within the 60-69 years age group.
+The highest number of patients are aged 60-64 years (1,006 patients), followed closely by 65-69 years (956 patients).
+The number of cases gradually decreases in older age groups, with 389 patients aged 80-84 years and 280 patients aged 85+ years.
+The youngest age groups (20-39 years) have significantly fewer cases, with only 3 patients aged 20-24 years and 64 patients aged 35-39 years.
+This distribution indicates that the majority of the study population consists of older adults, which aligns with typical cancer incidence patterns that increase with age.
+
+```{r}
+#| label: tbl-summaryage
+#| tbl-cap: "Summary of age."
+#| echo: FALSE
+resulttable=readRDS(here("results","edatables","edaagesummary.rds"))
+knitr::kable(resulttable)
+```
+
+```{r}
+#| label: fig-age 
+#| fig-cap: "Age"
+#| echo: FALSE
+knitr::include_graphics(here("results","edafigures","edaagedistribution.png"))
+```
+
+@tbl-summarymetastasis and @fig-mets shows distribution for different sit for metastasis.
+The table presents the distribution of metastatic sites at the time of diagnosis which is the main exposure in our study.
+Lung metastases were the most common, observed in 1,676 patients, followed by liver metastases (844 patients).
+Bone metastases were identified in 556 patients, while brain metastases were the least frequent, affecting 160 patients.
+This distribution suggests that lung and liver are the most frequent metastatic sites, which aligns with common metastatic patterns observed in many cancers.
+The lower occurrence of brain metastases may reflect differences in metastatic tropism or detection rates.
+
+```{r}
+#| label: tbl-summarymetastasis
+#| tbl-cap: "Summary of metastasis."
+#| echo: FALSE
+resulttable=readRDS(here("results","edatables","edametsatsummary.rds"))
+knitr::kable(resulttable)
+```
+
+```{r}
+#| label: fig-mets 
+#| fig-cap: "Metatstasis"
+#| echo: FALSE
+knitr::include_graphics(here("results","edafigures","edametsatdxdistribution.png"))
 ```
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
+*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
 simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
 
@@
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
+*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
 This might again be code that is best placed in one or several separate R scripts that need to be well documented.
 You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
+Then you load them here.*
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+
+*Summarize what you did, what you found and what it means.*
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+
+*Discuss what you perceive as strengths and limitations of your analysis.*
 
 ## Conclusions
-_What are the main take-home messages?_
 
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
+*What are the main take-home messages?*
+
+*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
 

### 2025-02-21 products/manuscript/Manuscript.qmd
part 2 project V2

@@
 However, this data specifically pertains to stage IV lung cancer patients.[@Mao2020] Patients with advanced endometrial cancer face poor survival outcomes, with metastatic disease being the leading cause of mortality.
-[@Amant2005] Therefore, this study aimed to utilize data from the Surveillance, Epidemiology, and End Results (SEER) database to analyze survival outcomes in patients diagnosed with stage IV endometrial cancer.
+[@Mao2020] Therefore, this study aimed to utilize data from the Surveillance, Epidemiology, and End Results (SEER) database to analyze survival outcomes in patients diagnosed with stage IV endometrial cancer.
 The primary objective was to determine whether the presence of lung/brain/bone/liver metastases in adults with stage IV endometrial cancer is associated with worse survival outcomes compared to patients without any metastasis.
@@
 To ensure completeness, individuals with missing or unknown survival months were excluded.
-The final cleaned dataset was saved for further analysis, ensuring consistency and reproducibility.The code and raw data to reproduce the data cleaning step is available in https://github.com/mohammedzuber099/ZUBER-MADA-project.
+The final cleaned dataset was saved for further analysis, ensuring consistency and reproducibility.
+The code and raw data to reproduce the data cleaning step is available in https://github.com/mohammedzuber099/ZUBER-MADA-project.
 
@@
 
+@fig-age shows this distribution indicates that the majority of the study population consists of older adults, which aligns with typical cancer incidence patterns that increase with age.
+
 ```{r}
@@
 Bone metastases were identified in 556 patients, while brain metastases were the least frequent, affecting 160 patients.
-This distribution suggests that lung and liver are the most frequent metastatic sites, which aligns with common metastatic patterns observed in many cancers.
-The lower occurrence of brain metastases may reflect differences in metastatic tropism or detection rates.
 
@@
 ```
+
+@fig-mets shows this distribution suggests that lung and liver are the most frequent metastatic sites, which aligns with common metastatic patterns observed in many cancers.
+The lower occurrence of brain metastases may reflect differences in metastatic tropism or detection rates.
 

### 2025-03-08 code/eda-code/edamz.qmd
Part 2 addressed issue2 based on Dr. @andreashandel comments. #2

@@
 saveRDS(summary_df, file = summarytable_file)
+```
+```{r}
+# Create a structured summary table
+var_summary <- mydata %>%
+  summarise_all(class) %>%
+  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Type") %>%
+  mutate(Unique_Values = sapply(mydata, function(x) length(unique(x))),
+         Missing_Values = sapply(mydata, function(x) sum(is.na(x))))
+
+# Print the table to verify
+print(var_summary)
+
+# Save the structured table in a readable format
+summarytable_file = here("results", "edatables", "edasummarytable2.rds")
+saveRDS(var_summary, file = summarytable_file)
 ```

### 2025-03-08 code/processing-code/processingcodemz.qmd
Part 2 addressed issue2 based on Dr. @andreashandel comments. #2

@@
 # data loading
+
 I am loading the raw data for further cleaning and processing
+
 ```{r}
@@
 ```
+
 # Check Data
@@
 Some variables were incorrectly read as numeric when they should be categorical (character).
-Convert the following:
-Year of diagnosis → Convert to character (since it represents a categorical year label rather than a numeric value).
+Convert the following: Year of diagnosis → Convert to character (since it represents a categorical year label rather than a numeric value).
 RX Summ--Surg Prim Site (1998+) → Convert to character (since it represents surgery codes rather than numerical values).
@@
 ```
+
 ## Converting Categorical Variables to Factor
 
-To ensure categorical variables are properly treated, we convert several key variables to factor type. This is important for:
-- Efficient storage and manipulation of categorical data.
+To ensure categorical variables are properly treated, we convert several key variables to factor type.
+This is important for: - Efficient storage and manipulation of categorical data.
 - Enabling statistical models and visualizations to properly recognize categories.
@@
 
-We apply this conversion to the following variables:
-- **Sex** and **Race-related variables** to ensure demographic information is handled correctly.
+We apply this conversion to the following variables: - **Sex** and **Race-related variables** to ensure demographic information is handled correctly.
 - **Cancer staging and metastasis indicators** to ensure proper classification.
 - **Cause of death and survival indicators** to maintain categorical consistency.
+
 ```{r}
@@
 ```
+
 ## Removing Patients in Specific Age Categories
 
-To focus our analysis on the adult patient population, we remove patients in the following age categories:
-- **01-04 years**
-- **15-19 years**
+To focus our analysis on the adult patient population, we remove patients in the following age categories: - **01-04 years** - **15-19 years**
 
@@
 ```
+
 ## Including Patients with Stage IVA and IVB
@@
 ```
+
 ## Excluding Patients with Specific SEER Cause-Specific Death Classification
 
-To ensure data quality, we exclude patients with `SEER cause-specific death classification` categorized as:
-- **Dead (missing/unknown COD)**
-- **N/A not seq 0-59**
+To ensure data quality, we exclude patients with `SEER cause-specific death classification` categorized as: - **Dead (missing/unknown COD)** - **N/A not seq 0-59**
 
@@
 ```
+
 ## All done, data is clean now.

### 2025-03-08 products/manuscript/Manuscript.qmd
Part 2 addressed issue2 based on Dr. @andreashandel comments. #2

@@
 #| echo: FALSE
-resulttable=readRDS(here("results","edatables","edasummarytable.rds"))
+resulttable=readRDS(here("results","edatables","edasummarytable2.rds"))
 knitr::kable(resulttable)

### 2025-03-14 code/analysis-code/statistical-analysis.qmd
part 3 project v1

---
title: "Statistical analysis"
format: html
editor: visual
---

This Quarto document presents the statistical data analysis for this project, focusing on modeling. It includes bivariate analyses to identify associations between key variables and a multivariate Cox proportional hazards model to assess survival outcomes. The analysis is conducted using the tidymodels framework and survival package in R.
# Loading Package
```{r}
#load needed packages. make sure they are installed.
# Load required packages. Ensure they are installed.
library(survival)  # For Cox regression analysis
library(survminer) # For survival curve visualization
library(broom)     # For tidying model output
library(tidyverse) # For data manipulation
library(here)      # For file path management (ensures reproducibility)
library(stringr)  # Ensure string manipulation functions are available
```
# Loadinf data
```{r}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")

#load data. 
mydata <- readRDS(data_location)
```
# Bivariate Analysis
The dataset is preprocessed to ensure consistency for survival analysis. First, all "Unknown" values in metastasis variables are removed, retaining only "Yes" and "No" categories. Next, metastasis variables are converted to factors with "No" as the reference category to ensure proper hazard ratio interpretation in Cox regression. A survival object is then created using "Survival months" as the time variable and "Vital status recode (study cutoff used)" to define event occurrence. These steps standardize the data for valid and reproducible survival analysis.

```{r}
# Remove "Unknown" from metastasis variables (keeping only "Yes" and "No")
mydata_filtered <- mydata %>%
  filter(mets_bone %in% c("Yes", "No"),
         mets_brain %in% c("Yes", "No"),
         mets_liver %in% c("Yes", "No"),
         mets_lung %in% c("Yes", "No"))

# Relevel metastasis variables so "No" is the reference category
mydata_filtered <- mydata_filtered %>%
  mutate(across(c(mets_bone, 
                  mets_brain, 
                  mets_liver, 
                  mets_lung), ~factor(.x, levels = c("No", "Yes"))))

# Create the survival object for Cox regression
surv_obj <- Surv(mydata_filtered$survival_months, mydata_filtered$vital_status == "Dead")

```

A Cox proportional hazards model is fitted separately for each metastasis variable to assess its impact on survival. The function dynamically constructs the regression formula and calculates hazard ratios (HRs) for "Yes" vs. "No" metastasis status. Results from all four models (bone, brain, liver, and lung metastases) are combined into a single table for better interpretation. The metastasis variable names are cleaned for clarity, and the final results are displayed and saved for reproducibility. This ensures a structured and systematic approach to analyzing the association between metastasis and survival outcomes.
## Bivariate Analysis: Metastasis and Survival
```{r}
####################################################
# Bivariate Analysis: Metastasis and Survival
####################################################
# Function to fit Cox regression for each metastasis variable
run_cox <- function(var) {
  formula <- as.formula(paste("surv_obj ~", paste0("`", var, "`")))  # Construct formula
  cox_model <- coxph(formula, data = mydata_filtered)  # Fit Cox model
  result <- broom::tidy(cox_model, exponentiate = TRUE)  # Convert log(HR) to HR
  return(result)  # Return model results
}

# Run Cox regression for each metastasis type (Yes vs. No comparison only)
cox_bone <- run_cox("mets_bone")
cox_brain <- run_cox("mets_brain")
cox_liver <- run_cox("mets_liver")
cox_lung <- run_cox("mets_lung")

# Combine the results into a single table
cox_results <- bind_rows(cox_bone, cox_brain, cox_liver, cox_lung)

# Rename terms for better readability
cox_results <- cox_results %>%
  mutate(term = case_when(
    term == "mets_boneYes"  ~ "Bone Metastasis",
    term == "mets_brainYes" ~ "Brain Metastasis",
    term == "mets_liverYes" ~ "Liver Metastasis",
    term == "mets_lungYes"  ~ "Lung Metastasis",
    TRUE ~ term  # Keep other terms unchanged
  ))

# Display the updated results
cox_results

# Save Cox results for reproducibility
saveRDS(cox_results, file = here::here("results", "tables", "mets_HR.rds"))

```
## KM curve Metastasis Bone
```{r}
 #Convert vital_status to a binary variable
mydata_filtered$survival_status <- ifelse(mydata_filtered$vital_status == "Dead", 1, 0)

# Create survival object
surv_obj <- Surv(mydata_filtered$survival_months, mydata_filtered$survival_status)

# Fit Kaplan-Meier survival curves for Bone Metastasis
km_fit <- survfit(surv_obj ~ mets_bone, data = mydata_filtered)

# Plot the Kaplan-Meier curves
km_plot <- ggsurvplot(
  km_fit,
  data = mydata_filtered,
  conf.int = TRUE,        # Show confidence intervals
  pval = TRUE,            # Display p-value
  risk.table = FALSE,      # Show risk table below the plot
  legend.title = "Bone Metastasis",
  legend.labs = c("No Bone Metastasis", "Bone Metastasis"),
  xlab = "Time (months)", # Adjust based on your dataset
  ylab = "Survival Probability",
  ggtheme = theme_minimal(),

)

# Print the KM plot
print(km_plot)

# Save the plot
ggsave(filename = here::here("results", "figures", "KM_curve_bone_mets.png"), plot = km_plot$plot, dpi = 300, width = 8, height = 6)

```
## KM curve Metastasis Brain
```{r}
 ####################################################
# Kaplan-Meier Analysis: Brain Metastasis and Survival
####################################################

# Load required libraries
library(survival)
library(survminer)

# Convert vital_status to a binary variable
mydata_filtered$survival_status <- ifelse(mydata_filtered$vital_status == "Dead", 1, 0)

# Create survival object
surv_obj <- Surv(mydata_filtered$survival_months, mydata_filtered$survival_status)

# Fit Kaplan-Meier survival curves for Brain Metastasis
km_fit <- survfit(surv_obj ~ mets_brain, data = mydata_filtered)

# Plot the Kaplan-Meier curves
km_plot <- ggsurvplot(
  km_fit,
  data = mydata_filtered,
  conf.int = TRUE,        # Show confidence intervals
  pval = TRUE,            # Display p-value
  risk.table = FALSE,     # Hide risk table (Set to TRUE if needed)
  legend.title = "Brain Metastasis",
  legend.labs = c("No Brain Metastasis", "Brain Metastasis"),
  xlab = "Time (months)", # Adjust based on your dataset
  ylab = "Survival Probability",
  ggtheme = theme_minimal()
)

# Print the KM plot
print(km_plot)

# Save the plot
ggsave(filename = here::here("results", "figures", "KM_curve_brain_mets.png"), plot = km_plot$plot, dpi = 300, width = 8, height = 6)

```
## KM curve Metastasis Liver
```{r}
####################################################
# Kaplan-Meier Analysis: Liver Metastasis and Survival
####################################################

# Load required libraries
library(survival)
library(survminer)

# Convert vital_status to a binary variable
mydata_filtered$survival_status <- ifelse(mydata_filtered$vital_status == "Dead", 1, 0)

# Create survival object
surv_obj <- Surv(mydata_filtered$survival_months, mydata_filtered$survival_status)

# Fit Kaplan-Meier survival curves for Liver Metastasis
km_fit <- survfit(surv_obj ~ mets_liver, data = mydata_filtered)

# Plot the Kaplan-Meier curves
km_plot <- ggsurvplot(
  km_fit,
  data = mydata_filtered,
  conf.int = TRUE,        # Show confidence intervals
  pval = TRUE,            # Display p-value
  risk.table = FALSE,     # Hide risk table (Set to TRUE if needed)
  legend.title = "Liver Metastasis",
  legend.labs = c("No Liver Metastasis", "Liver Metastasis"),
  xlab = "Time (months)", # Adjust based on your dataset
  ylab = "Survival Probability",
  ggtheme = theme_minimal()
)

# Print the KM plot
print(km_plot)

# Save the plot
ggsave(filename = here::here("results", "figures", "KM_curve_liver_mets.png"), plot = km_plot$plot, dpi = 300, width = 8, height = 6)

```

## KM curve Metastasis Lung 
```{r}
####################################################
# Kaplan-Meier Analysis: Lung Metastasis and Survival
####################################################

# Load required libraries
library(survival)
library(survminer)

# Convert vital_status to a binary variable
mydata_filtered$survival_status <- ifelse(mydata_filtered$vital_status == "Dead", 1, 0)

# Create survival object
surv_obj <- Surv(mydata_filtered$survival_months, mydata_filtered$survival_status)

# Fit Kaplan-Meier survival curves for Lung Metastasis
km_fit <- survfit(surv_obj ~ mets_lung, data = mydata_filtered)

# Plot the Kaplan-Meier curves
km_plot <- ggsurvplot(
  km_fit,
  data = mydata_filtered,
  conf.int = TRUE,        # Show confidence intervals
  pval = TRUE,            # Display p-value
  risk.table = FALSE,     # Hide risk table (Set to TRUE if needed)
  legend.title = "Lung Metastasis",
  legend.labs = c("No Lung Metastasis", "Lung Metastasis"),
  xlab = "Time (months)", # Adjust based on your dataset
  ylab = "Survival Probability",
  ggtheme = theme_minimal()
)

# Print the KM plot
print(km_plot)

# Save the plot
ggsave(filename = here::here("results", "figures", "KM_curve_lung_mets.png"), plot = km_plot$plot, dpi = 300, width = 8, height = 6)
```

## Bivariate Analysis: Race and Survival

The race variable is recoded as a factor with "White" as the reference category to ensure proper comparison in Cox regression. A separate model is fitted to assess the association between race and survival, with hazard ratios (HRs) calculated for Black, Asian or Pacific Islander, and American Indian/Alaska Native populations relative to White. The results are formatted for clarity, renaming the terms to indicate direct comparisons. Finally, the updated results are displayed and saved for reproducibility, allowing for structured and interpretable survival analysis by race.
```{r}
####################################################
# Bivariate Analysis: Race and Survival
####################################################

# Recode the race variable as a factor, ensuring "White" is the reference category
mydata_filtered <- mydata_filtered %>%
  mutate(race = factor(race, 
                                                 levels = c("White", "Black", "Asian or Pacific Islander", "American Indian/Alaska Native")))

# Function to fit Cox regression for Race categories
run_cox_race <- function() {
  # Construct Cox model formula using race variable
  formula <- as.formula("surv_obj ~ race")

  # Fit Cox model
  cox_model <- coxph(formula, data = mydata_filtered)

  # Convert output to tidy format and exponentiate coefficients to get Hazard Ratios (HRs)
  result <- broom::tidy(cox_model, exponentiate = TRUE)

  return(result)  # Return model results
}

# Run the Cox regression for Race categories
cox_race_results <- run_cox_race()

# Rename terms for better readability
cox_race_results <- cox_race_results %>%
  mutate(term = case_when(
    term == "raceBlack" ~ "Black vs. White",
    term == "raceAsian or Pacific Islander" ~ "Asian vs. White",
    term == "raceAmerican Indian/Alaska Native" ~ "American Indian vs. White",
    TRUE ~ term  # Keep other terms unchanged
  ))

# Display results
cox_race_results

# Save the results for reproducibility
saveRDS(cox_race_results, file = here::here("results", "tables", "race_HR.rds"))

```

## Kaplan-Meier Survival Curve: Race and Survival

```{r}
####################################################
# Kaplan-Meier Survival Curve: Race and Survival
####################################################

# Ensure survival variables are properly formatted
mydata_filtered <- mydata_filtered %>% 
  mutate(survival_months = as.numeric(survival_months),
         vital_status = ifelse(vital_status == "Dead", 1, 0))

# Handle NA values in survival_months
if(any(is.na(mydata_filtered$survival_months))) {
  cat("Warning: NAs introduced in survival_months due to coercion. Removing NAs for survival analysis.\n")
  mydata_filtered <- mydata_filtered %>% filter(!is.na(survival_months))
}

# Define survival object
surv_obj <- Surv(mydata_filtered$survival_months, mydata_filtered$vital_status)

# Fit Kaplan-Meier survival model for Race
fit_race <- survfit(surv_obj ~ race, data = mydata_filtered)

# Generate Kaplan-Meier plot for Race categories
ggsurvplot(fit_race, data = mydata_filtered, conf.int = TRUE, pval = TRUE, risk.table = TRUE, 
           title = "Kaplan-Meier Survival Curves by Race", 
           xlab = "Time in Months", ylab = "Survival Probability",
           legend.title = "Race",
           legend.labs = levels(mydata_filtered$race), # Dynamically assign race labels
           xlim = c(0, 130), ggtheme = theme_minimal())
# Save the plot
ggsave(filename = here::here("results", "figures", "KM_curve_race.png"), plot = km_plot$plot, dpi = 300, width = 8, height = 6)
```

## Bivariate Analysis: Age and Survival
The age variable is recoded as a factor with "20-24 years" set as the reference category to allow for meaningful comparisons in Cox regression. A separate model is fitted to evaluate the association between different age groups and survival outcomes, with hazard ratios (HRs) estimated relative to the 20-24 years category. The results are formatted to improve clarity, renaming each term to reflect direct age group comparisons. Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of age-related survival differences.

```{r}
####################################################
# Bivariate Analysis: Age and Survival
####################################################

# Function to fit Cox regression for Age groups
run_cox_age <- function() {
  # Construct Cox model formula using the updated age variable name
  formula <- as.formula("surv_obj ~ age_group")

  # Fit Cox model
  cox_model <- coxph(formula, data = mydata_filtered)

  # Convert output to tidy format and exponentiate coefficients to get Hazard Ratios (HRs)
  result <- broom::tidy(cox_model, exponentiate = TRUE)

  return(result)  # Return model results
}

# Run the Cox regression for Age categories
cox_age_results <- run_cox_age()

# Rename terms for better readability (comparing each group to 20-24 years)
cox_age_results <- cox_age_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    str_detect(term, "age_group25-29 years") ~ "25-29",
    str_detect(term, "age_group30-34 years") ~ "30-34",
    str_detect(term, "age_group35-39 years") ~ "35-39",
    str_detect(term, "age_group40-44 years") ~ "40-44",
    str_detect(term, "age_group45-49 years") ~ "45-49",
    str_detect(term, "age_group50-54 years") ~ "50-54",
    str_detect(term, "age_group55-59 years") ~ "55-59",
    str_detect(term, "age_group60-64 years") ~ "60-64",
    str_detect(term, "age_group65-69 years") ~ "65-69",
    str_detect(term, "age_group70-74 years") ~ "70-74",
    str_detect(term, "age_group75-79 years") ~ "75-79",
    str_detect(term, "age_group80-84 years") ~ "80-84",
    str_detect(term, "age_group85+ years") ~ "85+",
    TRUE ~ term  # Keep other terms unchanged
  ))

# Display results
cox_age_results

# Save the results for reproducibility
saveRDS(cox_age_results, file = here::here("results", "tables", "age_HR.rds"))

```
## Bivariate Analysis: Year of Diagnosis and Survival

```{r}
####################################################
# Bivariate Analysis: Year of Diagnosis and Survival
####################################################

# Identify the earliest year available in the dataset to set as reference
reference_year <- min(as.numeric(mydata_filtered$diagnosis_year), na.rm = TRUE)

# Convert Year of Diagnosis to a factor with the reference year as baseline
mydata_filtered <- mydata_filtered %>%
  mutate(diagnosis_year = factor(diagnosis_year, 
                                 levels = sort(unique(diagnosis_year), decreasing = FALSE)))

# Function to fit Cox regression for Year of Diagnosis
run_cox_year <- function() {
  # Construct Cox model formula using diagnosis_year variable
  formula <- as.formula("surv_obj ~ diagnosis_year")

  # Fit Cox model
  cox_model <- coxph(formula, data = mydata_filtered)

  # Convert output to tidy format and exponentiate coefficients to get Hazard Ratios (HRs)
  result <- broom::tidy(cox_model, exponentiate = TRUE)

  return(result)  # Return model results
}

# Run the Cox regression for Year of Diagnosis
cox_year_results <- run_cox_year()

# Rename terms for better readability (comparing each year to the reference year)
cox_year_results <- cox_year_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    TRUE ~ paste0(term, " vs. ", reference_year)  # Compare each year with the reference
  ))

# Display results
cox_year_results

# Save the results for reproducibility
saveRDS(cox_year_results, file = here::here("results", "tables", "year_HR.rds"))

```

## Bivariate Analysis: Median Household Income (Inflation-Adjusted to 2019)

The median household income variable, inflation-adjusted to 2019, is recoded as a factor with "$75,000+" as the reference category to allow for clear comparisons in Cox regression. A separate model is fitted to assess the association between income levels and survival outcomes, with hazard ratios (HRs) estimated relative to the highest income group. The results are formatted for better readability by appending comparison labels for each income group against "$75,000+." Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of socioeconomic disparities in survival.

```{r}
####################################################
# Bivariate Analysis: Median Household Income (Inflation-Adjusted to 2019)
####################################################
# Convert Household Income to a factor with "$75,000+" as the reference category
mydata_filtered <- mydata_filtered %>%
  mutate(median_income = factor(median_income, 
                                                                  levels = c("$75,000+", 
                                                                             "$35,000 - $39,999", 
                                                                             "$40,000 - $44,999", 
                                                                             "$45,000 - $49,999", 
                                                                             "$50,000 - $54,999",
                                                                             "$55,000 - $59,999",
                                                                             "$60,000 - $64,999",
                                                                             "$65,000 - $69,999",
                                                                             "$70,000 - $74,999")))

# Function to fit Cox regression for Household Income
run_cox_income <- function() {
  # Construct Cox model formula using Household Income variable
  formula <- as.formula("surv_obj ~ median_income")

  # Fit Cox model
  cox_model <- coxph(formula, data = mydata_filtered)

  # Convert output to tidy format and exponentiate coefficients to get Hazard Ratios (HRs)
  result <- broom::tidy(cox_model, exponentiate = TRUE)

  # Add column for variable reference
  result$Variable <- "Household Income"

  return(result)  # Return model results
}

# Run the Cox regression for Household Income
cox_income_results <- run_cox_income()

# Rename terms for better readability (comparing each income group to "$75,000+")
cox_income_results <- cox_income_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    TRUE ~ paste0(term, " vs. $75,000+")
  ))

# Display results
cox_income_results

# Save the results for reproducibility
saveRDS(cox_income_results, file = here::here("results", "tables", "income_HR_75K.rds"))

```

# Multivariable Cox Proportional Hazards Model
```{r}
####################################################
# Multivariable Cox Proportional Hazards Model
####################################################

# Define the multivariable Cox regression formula with all selected covariates
multivariable_formula <- as.formula("surv_obj ~ 
  mets_bone + 
  mets_brain + 
  mets_liver + 
  mets_lung + 
  race + 
  age_group + 
  diagnosis_year + 
  median_income")

# Fit the Cox proportional hazards model
multivariable_cox <- coxph(multivariable_formula, data = mydata_filtered)

# Convert model output to a tidy format and exponentiate coefficients to get Hazard Ratios (HRs)
multivariable_results <- broom::tidy(multivariable_cox, exponentiate = TRUE)

# Rename terms for better readability in the output
multivariable_results <- multivariable_results %>%
  mutate(term = str_replace_all(term, "`", "")) %>%
  mutate(term = case_when(
    term == "mets_boneYes" ~ "Bone Metastasis",
    term == "mets_brainYes" ~ "Brain Metastasis",
    term == "mets_liverYes" ~ "Liver Metastasis",
    term == "mets_lungYes" ~ "Lung Metastasis",
    term == "raceBlack" ~ "Black",
    term == "raceAsian or Pacific Islander" ~ "Asian or Pacific Islander",
    term == "raceAmerican Indian/Alaska Native" ~ "American Indian/Alaska Native",
    str_detect(term, "age_group") ~ str_remove(term, "age_group") %>% str_remove(" vs. 20-24 years"),
    str_detect(term, "household_income") ~ str_remove(term, "household_income") %>% str_remove(" vs. \\$75,000\\+"),
    TRUE ~ term  # Keep other terms unchanged
  ))

# Display the multivariable Cox model results
multivariable_results

# Save the results for reproducibility
saveRDS(multivariable_results, file = here::here("results", "tables", "multivariable_cox_HR.rds"))

```

### 2025-03-14 code/eda-code/edamz.qmd
part 3 project v1

@@
 p1 <- mydata %>%
-  group_by(`Age recode with <1 year olds`) %>%
+  group_by(`age_group`) %>%
   summarise(Count = n())
@@
 p2 <- mydata %>%
-  ggplot(aes(x = `Age recode with <1 year olds`)) + 
+  ggplot(aes(x = `age_group`)) + 
   geom_bar(fill = "skyblue") + 
@@
 p3 <- mydata %>%
-  group_by(`Race recode (W, B, AI, API)`) %>%
+  group_by(`race`) %>%
   summarise(Count = n())
@@
 p4 <- mydata %>%
-  ggplot(aes(x = `Race recode (W, B, AI, API)`)) + 
+  ggplot(aes(x = race)) + 
   geom_bar(fill = "skyblue") + 
@@
 p5 <- mydata %>%
-  group_by(`Year of diagnosis`) %>%
+  group_by(diagnosis_year) %>%
   summarise(Count = n())
@@
 p6 <- mydata %>%
-  ggplot(aes(x = factor(`Year of diagnosis`))) +  # Treat as categorical
+  ggplot(aes(x = factor(diagnosis_year))) +  # Treat as categorical
   geom_bar(fill = "skyblue") + 
@@
   summarise(
-    Bone_Mets = sum(`SEER Combined Mets at DX-bone (2010+)` == "Yes", na.rm = TRUE),
-    Brain_Mets = sum(`SEER Combined Mets at DX-brain (2010+)` == "Yes", na.rm = TRUE),
-    Liver_Mets = sum(`SEER Combined Mets at DX-liver (2010+)` == "Yes", na.rm = TRUE),
-    Lung_Mets = sum(`SEER Combined Mets at DX-lung (2010+)` == "Yes", na.rm = TRUE),
+    Bone_Mets = sum(mets_bone == "Yes", na.rm = TRUE),
+    Brain_Mets = sum(mets_brain == "Yes", na.rm = TRUE),
+    Liver_Mets = sum(mets_liver == "Yes", na.rm = TRUE),
+    Lung_Mets = sum(mets_lung == "Yes", na.rm = TRUE),
       ) %>%
@@
 p9 <- mydata %>%
-  group_by(`SEER cause-specific death classification`) %>%
+  group_by(cause_specific_death) %>%
   summarise(Count = n(), .groups = "drop") %>%
@@
 p10 <- p9 %>%
-  ggplot(aes(x = `SEER cause-specific death classification`, y = Count, fill = `SEER cause-specific death classification`)) + 
+  ggplot(aes(x =cause_specific_death, y = Count, fill = cause_specific_death)) + 
   geom_bar(stat = "identity") + 
@@
   summarise(
-    Min_Survival = min(`Survival months`, na.rm = TRUE),
-    Max_Survival = max(`Survival months`, na.rm = TRUE),
-    Mean_Survival = mean(`Survival months`, na.rm = TRUE),
-    Median_Survival = median(`Survival months`, na.rm = TRUE),
-    SD_Survival = sd(`Survival months`, na.rm = TRUE),
+    Min_Survival = min(survival_months, na.rm = TRUE),
+    Max_Survival = max(survival_months, na.rm = TRUE),
+    Mean_Survival = mean(survival_months, na.rm = TRUE),
+    Median_Survival = median(survival_months, na.rm = TRUE),
+    SD_Survival = sd(survival_months, na.rm = TRUE),
     Count = n()
@@
 p12 <- mydata %>%
-  ggplot(aes(x = `Survival months`)) + 
+  ggplot(aes(x = survival_months)) + 
   geom_histogram(binwidth = 5, fill = "skyblue", color = "black") + 

### 2025-03-14 code/processing-code/processingcodemz.qmd
part 3 project v1

@@
 
+## Excluding Patients with Unknown metastasis
+
+```{r}
+d7 <- d6 %>% dplyr::filter(
+    # Exclude patients where all four metastasis indicators are "Unknown"
+    !(`SEER Combined Mets at DX-bone (2010+)` == "Unknown" &
+      `SEER Combined Mets at DX-brain (2010+)` == "Unknown" &
+      `SEER Combined Mets at DX-liver (2010+)` == "Unknown" &
+      `SEER Combined Mets at DX-lung (2010+)` == "Unknown")
+)
+
+# Check the updated dataset after filtering
+skimr::skim(d7)
+
+```
+
+## Renaming all the variables for the ease of analysis 
+```{r}
+d8 <- d7 %>% dplyr::rename(
+    # Demographic Information
+    age_group = `Age recode with <1 year olds`,  # Age category of the patient
+    sex = Sex,  # Patient's sex
+    race = `Race recode (W, B, AI, API)`,  # Race classification
+    ethnicity = `Origin recode NHIA (Hispanic, Non-Hisp)`,  # Ethnicity classification
+    race_ethnicity = `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`,  # Combined race & ethnicity
+
+    # Diagnosis & Cancer Staging
+    diagnosis_year = `Year of diagnosis`,  # Year of cancer diagnosis
+    cancer_stage = `Derived AJCC Stage Group, 7th ed (2010-2015)`,  # Cancer staging classification
+    surgery_primary_site = `RX Summ--Surg Prim Site (1998+)`,  # Surgery performed at the primary site
+
+    # Metastasis Indicators
+    mets_bone = `SEER Combined Mets at DX-bone (2010+)`,  # Metastasis at bone
+    mets_brain = `SEER Combined Mets at DX-brain (2010+)`,  # Metastasis at brain
+    mets_liver = `SEER Combined Mets at DX-liver (2010+)`,  # Metastasis at liver
+    mets_lung = `SEER Combined Mets at DX-lung (2010+)`,  # Metastasis at lung
+    mets_distant_ln = `Mets at DX-Distant LN (2016+)`,  # Metastasis at distant lymph nodes
+
+    # Cause of Death & Survival
+    cause_of_death = `COD to site recode`,  # Cause of death classification
+    cause_specific_death = `SEER cause-specific death classification`,  # SEER-specific cause of death
+    other_cause_of_death = `SEER other cause of death classification`,  # Other causes of death
+    survival_months = `Survival months`,  # Number of survival months
+    survival_flag = `Survival months flag`,  # Flag indicating survival data availability
+    vital_status = `Vital status recode (study cutoff used)`,  # Vital status (Alive/Dead)
+    cod_km = `COD to site rec KM`,  # Cause of death classification for Kaplan-Meier
+
+    # Socioeconomic & Patient ID
+    patient_id = `Patient ID`,  # Unique patient identifier
+    median_income = `Median household income inflation adj to 2019`,  # Adjusted median income
+    rural_urban_code = `Rural-Urban Continuum Code`  # Rural-Urban classification
+)
+
+# Check new variable names to ensure proper renaming
+colnames(d8)
+
+```
+
 ## All done, data is clean now.
@@
 ```{r}
-processeddata <- d6  # Assign cleaned dataset
+processeddata <- d8  # Assign cleaned dataset
 ```
@@
 ```
-
-# Notes
-
-Removing individuals with "faulty" or missing data is one approach.
-However, depending on the question and analysis approach, different cleaning methods might be better (e.g., keeping individuals with some missing information).

### 2025-03-14 products/manuscript/Manuscript.qmd
part 3 project v1

@@
 ---
-title: "The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer:A SEER-Based Cohort Study"
+title: "The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort Study"
 author: "Mohammed Zuber"
@@
 ---
-
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 The incidence of metastasis will be calculated in stage IV endometrial cancer.
-A Cox regression model will be performed to compute the adjusted Hazard Ratios (aHR) and 95% confidence intervals (CI).
+In the bivariate analysis, we fitted separate Cox regression models for each variable, including metastasis status (bone, brain, liver, and lung), race, age group, year of diagnosis, and median household income, to estimate their individual hazard ratios (HRs).
+For the multivariate analysis, we constructed a Cox model incorporating all covariates simultaneously to adjust for potential confounders and determine independent predictors of survival.
+A Cox regression model was performed to compute the adjusted Hazard Ratios (aHR).
 
@@
 ## Basic statistical analysis
+### Bivariate Analysis: Metastasis and Survival
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
+In the bivariate Cox proportional hazards analysis, all four metastasis types were significantly associated with an increased risk of mortality.
+Bone metastasis was associated with a 56% higher hazard of death (HR = 1.56), while brain metastasis showed the highest risk, nearly a 92% increased hazard of death (HR = 1.92).
+Liver metastasis was also strongly associated with worse survival, showing an 81% higher hazard of death (HR = 1.81), and lung metastasis increased the hazard of death by 63% (HR = 1.63).
+These findings indicate that the presence of metastases, particularly in the brain and liver, is a strong predictor of poor survival outcomes.
+@tbl-summaryhrmetastasis shows the HR for different metastatis sites
 
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
+#| label: tbl-summaryhrmetastasis
+#| tbl-cap: "Proportional hazards analysis for different metastasis."
 #| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
+resulttable=readRDS(here("results", "tables", "mets_HR.rds"))
+knitr::kable(resulttable)
 ```
+
+@fig-km-curve-bone-mets illustrates the survival probability of patients with and without bone metastasis over time.
+The red curve represents patients without bone metastasis, while the blue curve represents those with bone metastasis.
+Patients with bone metastasis have significantly worse survival outcomes, as indicated by the lower survival probability throughout the follow-up period.
+
+```{r}
+#| label: fig-km-curve-bone-mets
+#| fig-cap: "Kaplan-Meier survival curves for Bone Metastasis"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "KM_curve_bone_mets.png"))
+```
+
+@fig-km-curve-brain-mets shows that patients with brain metastasis have significantly lower survival probabilities than those without.
+The log-rank test (p \< 0.0001) confirms a statistically significant difference, indicating that brain metastasis is associated with worse survival outcomes.
+
+```{r}
+#| label: fig-km-curve-brain-mets
+#| fig-cap: "Kaplan-Meier survival curves for Brain Metastasis"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "KM_curve_brain_mets.png"))
+```
+
+@fig-km-curve-liver-mets indicates that patients with liver metastasis (blue) have significantly lower survival probabilities than those without (red).
+The log-rank test (p \< 0.0001) confirms a statistically significant difference, suggesting that liver metastasis is associated with poorer survival outcomes.
+
+```{r}
+#| label: fig-km-curve-liver-mets
+#| fig-cap: "Kaplan-Meier survival curves for Liver Metastasis"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "KM_curve_liver_mets.png"))
+```
+
+@fig-km-curve-lung-mets shows that patients with lung metastasis (blue) have significantly lower survival probabilities than those without (red).
+The log-rank test (p \< 0.0001) confirms a statistically significant difference, indicating that lung metastasis is associated with worse survival outcomes.
+
+```{r}
+#| label: fig-km-curve-lung-mets
+#| fig-cap: "Kaplan-Meier survival curves for Lung Metastasis"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "KM_curve_lung_mets.png"))
+```
+### Bivariate Analysis: Race and Survival
+Patients identified as  had significantly worse survival outcomes than White patients, with a 37% higher risk of death (HR = 1.37, p < 0.0001).
+In contrast, Asian patients had a lower risk of death compared to White patients (HR = 0.86, p = 0.0076), suggesting a potential survival advantage.
+The survival difference between American Indian and White patients was not statistically significant (HR = 1.21, p = 0.29), indicating no clear evidence of an increased or decreased risk.
+@tbl-summaryhrrace shows the HR for patients with different race.
+```{r}
+#| label: tbl-summaryhrrace
+#| tbl-cap: "Proportional hazards analysis for different Race."
+#| echo: FALSE
+resulttable=readRDS(here("results", "tables", "race_HR.rds"))
+knitr::kable(resulttable)
+```
+### Bivariate Analysis: Age and Survival
+The hazard ratio (HR) estimates for different age groups indicate a general trend of increasing risk with age, though statistical significance varies.
+Compared to the reference group (20-24 years), younger age groups (25-74 years) show HRs between 3.5 and 6.2, but their p-values are above 0.05, suggesting no statistically significant difference in survival risk.
+However, individuals aged 80-84 years (HR = 7.99, p = 0.038) and 85+ years (HR = 11.01, p = 0.017) have significantly higher mortality risk, indicating poorer survival outcomes in older patients.
+These findings suggest that age is an important predictor of survival, with the highest risk observed in the oldest age group.
+@tbl-summaryhrage shows the HR for diferrent age groups.
+```{r}
+#| label: tbl-summaryhrage
+#| tbl-cap: "Proportional hazards analysis for different age group."
+#| echo: FALSE
+resulttable=readRDS(here("results", "tables", "age_HR.rds"))
+knitr::kable(resulttable)
+```
+### Multivariable analysis
+
+The multivariable Cox proportional hazards model results indicate that metastasis to the bone (HR = 1.30, p < 0.001), brain (HR = 1.72, p < 0.001), liver (HR = 1.66, p < 0.001), and lung (HR = 1.51, p < 0.001) are all significantly associated with poorer survival outcomes.
+Among racial groups, Black patients (HR = 1.39, p < 0.001) had significantly worse survival compared to White patients, while Asian or Pacific Islander and American Indian/Alaska Native groups did not show a statistically significant difference.
+
+Regarding age, survival risk generally increased with older age, with statistically significant increases observed in the 70-74 years (HR = 7.26, p = 0.048), 75-79 years (HR = 7.91, p = 0.039), 80-84 years (HR = 9.53, p = 0.025), and 85+ years (HR = 12.75, p = 0.011) groups, indicating poorer survival outcomes in older patients.
+
+Year of diagnosis did not show a consistent or significant trend, suggesting no clear improvement or decline in survival based on the year of diagnosis.
+Household income was associated with survival, with patients in the $50,000 - $54,999 (HR = 1.16, p = 0.011) and $65,000 - $69,999 (HR = 1.12, p = 0.030) categories showing significantly higher risk compared to those earning $75,000+.
+However, other income groups did not show significant differences.
+
+Overall, metastasis, age, and race appear to be key determinants of survival, with Black race and older age groups associated with poorer outcomes.
 

### 2025-04-03 code/analysis-code/statistical-analysis.qmd
part 4 project

@@
 The analysis is conducted using the tidymodels framework and survival package in R.
-# Loading Package
+\# Loading Package
+
 ```{r}
@@
 library(stringr)  # Ensure string manipulation functions are available
+library(tidymodels)
+library(vip)
+library(themis)
+library(randomForestSRC)
+library(dplyr)
+library(ranger)
+library(survival)
+library(pROC)
+library(caret)
+library(glmnet)
 ```
-# Loadinf data
+
+# Loading data
+
 ```{r}
@@
 ```
+
 # Bivariate Analysis
+
 The dataset is preprocessed to ensure consistency for survival analysis.
@@
 This ensures a structured and systematic approach to analyzing the association between metastasis and survival outcomes.
-## Bivariate Analysis: Metastasis and Survival
+\## Bivariate Analysis: Metastasis and Survival
+
 ```{r}
@@
 ```
+
 ## KM curve Metastasis Bone
+
 ```{r}
@@
 ```
+
 ## KM curve Metastasis Brain
+
 ```{r}
@@
 ```
+
 ## KM curve Metastasis Liver
+
 ```{r}
@@
 ## KM curve Metastasis Lung
+
 ```{r}
@@
 Finally, the updated results are displayed and saved for reproducibility, allowing for structured and interpretable survival analysis by race.
+
 ```{r}
@@
 ## Bivariate Analysis: Age and Survival
+
 The age variable is recoded as a factor with "20-24 years" set as the reference category to allow for meaningful comparisons in Cox regression.
@@
 ```
+
 ## Bivariate Analysis: Year of Diagnosis and Survival
@@
 
-The median household income variable, inflation-adjusted to 2019, is recoded as a factor with "$75,000+" as the reference category to allow for clear comparisons in Cox regression.
+The median household income variable, inflation-adjusted to 2019, is recoded as a factor with "\$75,000+" as the reference category to allow for clear comparisons in Cox regression.
 A separate model is fitted to assess the association between income levels and survival outcomes, with hazard ratios (HRs) estimated relative to the highest income group.
-The results are formatted for better readability by appending comparison labels for each income group against "$75,000+." Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of socioeconomic disparities in survival.
+The results are formatted for better readability by appending comparison labels for each income group against "\$75,000+." Finally, the updated results are displayed and saved for reproducibility, ensuring a structured and interpretable analysis of socioeconomic disparities in survival.
 
@@
 # Multivariable Cox Proportional Hazards Model
+
 ```{r}
@@
 ```
+
+# Machine Learning Model
+
+I used the random forest for predicting cause specifc death but the model did not perform well. Therefore, using the random survival forest model.
+
+A random forest classification model was developed to predict cause-specific death using demographic and clinical features, including metastasis status, age group, race, income, and year of diagnosis.
+The model was trained on 80% of the data and evaluated on a 20% test set, with hyperparameter tuning via 5-fold cross-validation.
+Despite identifying several important predictors—most notably lung and liver metastases, age group 45–49, bone metastasis, and race—the model demonstrated poor overall predictive performance.
+The area under the ROC curve (AUC) was 0.39, indicating that the model performed only marginally better than random chance in distinguishing between patients who died and those who survived.
+The confusion matrix further reflected this limitation, showing moderate sensitivity but poor specificity.
+These results suggest that while random forest can highlight relevant variables, it is not well-suited for modeling survival outcomes in the presence of censoring.
+In contrast, approaches such as random survival forests (RSF) or Cox-based models offer more reliable and interpretable performance in survival analysis contexts.
+\## Random Forest
+
+```{r}
+####################################################
+# Random Forest with Class Balancing (Downsampling)
+# Goal: Predict cause-specific death using SEER variables
+####################################################
+
+# Step 1: Prepare the dataset
+rf_data <- mydata_filtered %>%
+  select(cause_specific_death, age_group, race, median_income, 
+         mets_bone, mets_brain, mets_liver, mets_lung, diagnosis_year) %>%
+  drop_na()
+
+# Step 2: Recode outcome to binary factor ("Yes"/"No")
+rf_data <- rf_data %>%
+  mutate(cause_specific_death = case_when(
+    cause_specific_death == "Dead (attributable to this cancer dx)" ~ "Yes",
+    cause_specific_death == "Alive or dead of other cause" ~ "No",
+    TRUE ~ NA_character_
+  )) %>%
+  filter(!is.na(cause_specific_death)) %>%
+  mutate(cause_specific_death = factor(cause_specific_death, levels = c("No", "Yes")))
+
+# Step 3: Split the data (80% training / 20% testing), stratified by class
+set.seed(123)
+rf_split <- initial_split(rf_data, prop = 0.8, strata = cause_specific_death)
+rf_train <- training(rf_split)
+rf_test  <- testing(rf_split)
+
+# Step 4: Preprocessing recipe with downsampling for class balance
+rf_recipe <- recipe(cause_specific_death ~ ., data = rf_train) %>%
+  step_dummy(all_nominal_predictors()) %>%
+  step_zv(all_predictors()) %>%
+  step_downsample(cause_specific_death)
+
+# Step 5: Define random forest model (with hyperparameter tuning)
+rf_model <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) %>%
+  set_engine("ranger", importance = "impurity") %>%
+  set_mode("classification")
+
+# Step 6: Create workflow
+rf_workflow <- workflow() %>%
+  add_recipe(rf_recipe) %>%
+  add_model(rf_model)
+
+# Step 7: 5-fold cross-validation
+set.seed(123)
+rf_folds <- vfold_cv(rf_train, v = 5)
+
+# Step 8: Tune the model
+rf_tuned <- tune_grid(
+  rf_workflow,
+  resamples = rf_folds,
+  grid = 10,
+  metrics = metric_set(roc_auc, accuracy)
+)
+
+# Step 9: Select best model based on ROC AUC
+best_rf <- select_best(rf_tuned, metric = "roc_auc")
+
+# Step 10: Finalize and train final model
+final_rf <- finalize_workflow(rf_workflow, best_rf)
+rf_fit <- fit(final_rf, data = rf_train)
+
+# Step 11: Make predictions on test set
+rf_predictions <- predict(rf_fit, rf_test, type = "prob") %>%
+  bind_cols(predict(rf_fit, rf_test)) %>%
+  bind_cols(rf_test)
+
+# Step 12: Evaluate model
+rf_auc <- roc_auc(rf_predictions, truth = cause_specific_death, .pred_Yes)
+rf_cm <- conf_mat(rf_predictions, truth = cause_specific_death, estimate = .pred_class)
+
+# Step 13: Variable importance plot
+vip_plot <- rf_fit %>%
+  extract_fit_parsnip() %>%
+  vip(num_features = 10)
+
+# Step 14: Print results
+rf_auc
+rf_cm
+vip_plot
+
+# Step 15: Save outputs
+saveRDS(rf_fit, here("results", "tables", "random_forest_balanced_model.rds"))
+saveRDS(rf_auc, here("results", "tables", "random_forest_balanced_auc.rds"))
+saveRDS(rf_cm, here("results", "tables", "random_forest_balanced_confusion_matrix.rds"))
+ggsave(filename = here("results", "figures", "vip_plot_rf_balanced.png"), plot = vip_plot, dpi = 300, width = 8, height = 6)
+```
+
+## Random Survival Forest
+
+A random survival forest model was trained on clinical and demographic features to predict survival outcomes.
+Variable importance analysis revealed that age group, median household income, diagnosis year, and race were the top contributors to survival prediction, whereas metastasis variables (bone, lung, liver, brain) were less influential.
+The model demonstrated moderate performance, with an AUC of \~0.72 at 12 months.
+While sensitivity was high (86.5%), specificity remained modest (47.9%), indicating that the model more accurately identified patients who survived.
+Individual-level survival curves were generated, illustrating patient-specific risk trajectories.
+
+```{r}
+####################################################
+# Random Survival Forest (RSF)
+# Goal: Model survival time using censored data
+####################################################
+
+# Step 1: Prepare and clean data
+rsf_data <- mydata_filtered %>%
+  mutate(event = vital_status) %>%
+  select(survival_months, event,
+         mets_lung, mets_liver, mets_bone, mets_brain,
+         race, age_group, median_income, diagnosis_year) %>%
+  # Drop rows with any NA in predictors or outcome
+  filter(if_all(everything(), ~ !is.na(.))) %>%
+  mutate(
+    mets_lung = as.factor(mets_lung),
+    mets_liver = as.factor(mets_liver),
+    mets_bone = as.factor(mets_bone),
+    mets_brain = as.factor(mets_brain),
+    race = as.factor(race),
+    age_group = as.factor(age_group),
+    median_income = as.factor(median_income),
+    diagnosis_year = as.factor(diagnosis_year)
+  )
+
+# Step 2: Create survival object
+rsf_data$surv_obj <- with(rsf_data, Surv(survival_months, event))
+
+# Step 3: Fit survival random forest using ranger
+rsf_model <- ranger(
+  formula = surv_obj ~ mets_lung + mets_liver + mets_bone + mets_brain +
+    race + age_group + median_income + diagnosis_year,
+  data = rsf_data,
+  num.trees = 300,
+  mtry = 3,
+  splitrule = "logrank",  # For survival
+  importance = "impurity",
+  seed = 123
+)
+
+# Step 4: Plot variable importance
+vip(rsf_model)
+
+```
+
+```{r}
+####################################################
+# RSF Results: Top Variables, Survival Prediction, AUC & Confusion Matrix
+####################################################
+# --- Step 1: Top Variable Importance ---
+importance_df <- data.frame(
+  variable = names(rsf_model$variable.importance),
+  importance = rsf_model$variable.importance
+) %>%
+  arrange(desc(importance))
+
+# View top 10
+head(importance_df, 10)
+
+# Save top 10
+write.csv(head(importance_df, 10), here::here("results", "tables", "top_10_rsf_variables.csv"), row.names = FALSE)
+
+# Plot top 10 variables
+vip::vip(rsf_model, num_features = 10)
+
+# --- Step 2: Predict Survival Curves ---
+# Predict full survival curve for all patients
+surv_preds <- predict(rsf_model, data = rsf_data, type = "response")
+
+# Plot survival curve for the first patient
+plot(rsf_model$unique.death.times, surv_preds$survival[1, ], type = "l",
+     xlab = "Time (months)", ylab = "Survival Probability",
+     main = "Survival Curve for Patient 1")
+
+# --- Step 3: Classify Patients Based on Survival at 12 Months ---
+# Choose cutoff time point (e.g., 12 months)
+target_time <- 12
+closest_index <- which.min(abs(rsf_model$unique.death.times - target_time))
+
+# Predicted survival prob at 12 months
+pred_surv_12 <- surv_preds$survival[, closest_index]
+
+# Convert to binary prediction: 1 = died before 12 months (low survival prob)
+pred_class <- ifelse(pred_surv_12 < 0.5, 1, 0)
+true_class <- rsf_data$event  # Actual: 1 = dead, 0 = alive
+
+# --- Step 4: Confusion Matrix ---
+conf_matrix <- caret::confusionMatrix(factor(pred_class), factor(true_class))
+print(conf_matrix)
+
+# --- Step 5: ROC Curve and AUC ---
+roc_obj <- roc(response = true_class, predictor = 1 - pred_surv_12)
+auc_value <- auc(roc_obj)
+print(paste("AUC at 12 months:", round(auc_value, 3)))
+
+# Plot ROC
+plot(roc_obj, main = "ROC Curve at 12 Months", col = "blue", lwd = 2)
+
+# --- Step 6: Save Outputs ---
+# Create and save top variable importance plot
+importance_plot <- vip::vip(rsf_model, num_features = 10)
+
+ggsave(
+  filename = here::here("results", "figures", "rsf_variable_importance.png"),
+  plot = importance_plot,
+  width = 8,
+  height = 6,
+  dpi = 300
+)
+
+# Save survival curve plot using base R
+png(filename = here::here("results", "figures", "rsf_survival_curve_patient1.png"),
+    width = 800, height = 600)
+
+plot(rsf_model$unique.death.times, surv_preds$survival[1, ], type = "l",
+     xlab = "Time (months)", ylab = "Survival Probability",
+     main = "Survival Curve for Patient 1", col = "darkgreen", lwd = 2)
+
+dev.off()
+
+# Save ROC curve plot using base R
+png(filename = here::here("results", "figures", "rsf_auc_12mo.png"),
+    width = 800, height = 600)
+
+plot(roc_obj, main = "ROC Curve at 12 Months", col = "blue", lwd = 2)
+dev.off()
+
+# Save confusion matrix text
+sink(file = here::here("results", "tables", "rsf_confusion_matrix.txt"))
+print(conf_matrix)
+sink()
+
+# Optional: Save a visual confusion matrix (if you have `autoplot` or ggplot-based CM)
+# e.g., using fourfoldplot for base R
+png(filename = here::here("results", "figures", "rsf_conf_matrix_12mo.png"),
+    width = 800, height = 600)
+
+fourfoldplot(conf_matrix$table, color = c("#E41A1C", "#377EB8"),
+             main = "RSF Confusion Matrix (12 Months)")
+
+dev.off()
+
+```
+
+## Lasso-Cox Model
+
+A LASSO-penalized Cox proportional hazards model was applied to the survival dataset to perform variable selection and reduce model complexity.
+Cross-validation identified the optimal penalty parameter (λ) that minimized the partial likelihood deviance.
+The model selected 10 predictors with non-zero coefficients, including all four metastasis variables (mets_lung, mets_liver, mets_bone, mets_brain) and several younger age groups (with negative coefficients, indicating lower risk).
+The results support the prognostic value of metastasis presence and younger age in overall survival prediction.
+
+```{r}
+####################################################
+# LASSO Cox Regression using glmnet
+####################################################
+# Step 1: Prepare data
+# Convert all categorical variables to dummy variables
+# Step 1: Prepare data
+lasso_data <- rsf_data %>%
+  select(survival_months, event, mets_lung, mets_liver, mets_bone, mets_brain,
+         age_group, race, median_income, diagnosis_year) %>%
+  filter(if_all(everything(), ~ !is.na(.))) %>%      # Remove rows with any NA
+  filter(survival_months > 0)                        # Remove 0 or negative times
+
+# Create model matrix (dummy coding all predictors)
+x <- model.matrix(~ mets_lung + mets_liver + mets_bone + mets_brain +
+                    age_group + race + median_income + diagnosis_year,
+                  data = lasso_data)[, -1]  # drop intercept
+
+# Create survival object
+y <- Surv(lasso_data$survival_months, lasso_data$event)
+
+# Step 2: Run cross-validated LASSO Cox
+set.seed(123)
+cvfit <- cv.glmnet(x, y, family = "cox", alpha = 1)
+
+# Plot cross-validation curve
+plot(cvfit)
+
+# Step 3: Extract best lambda and coefficients
+best_lambda <- cvfit$lambda.min
+lasso_coef <- coef(cvfit, s = "lambda.min")
+
+# Convert to tidy format
+lasso_selected <- as.data.frame(as.matrix(lasso_coef))
+lasso_selected <- lasso_selected %>%
+  tibble::rownames_to_column("variable") %>%
+  rename(coefficient = `1`) %>%
+  filter(coefficient != 0)
+
+# Step 4: View selected variables
+print(lasso_selected)
+
+# Optional: Save selected variables
+write.csv(lasso_selected, "results/tables/lasso_selected_variables.csv", row.names = FALSE)
+
+```
+
+```{r}
+####################################################
+# LASSO Cox: AUC and Confusion Matrix at 12 Months
+####################################################
+
+# Step 1: Use same cleaned dataset
+# Make sure this matches the data used to train LASSO
+lasso_data <- rsf_data %>%
+  select(survival_months, event, mets_lung, mets_liver, mets_bone, mets_brain,
+         age_group, race, median_income, diagnosis_year) %>%
+  filter(if_all(everything(), ~ !is.na(.))) %>%
+  filter(survival_months > 0)
+
+# Model matrix
+x <- model.matrix(~ mets_lung + mets_liver + mets_bone + mets_brain +
+                    age_group + race + median_income + diagnosis_year,
+                  data = lasso_data)[, -1]
+
+# Step 2: Predict linear risk scores
+risk_scores <- predict(cvfit, newx = x, s = "lambda.min", type = "link")
+
+# Step 3: Define binary outcome at 12 months
+target_time <- 12
+true_class <- ifelse(lasso_data$survival_months <= target_time & lasso_data$event == 1, 1, 0)
+
+# Step 4: AUC
+roc_obj <- roc(response = true_class, predictor = as.vector(risk_scores))
+auc_val <- auc(roc_obj)
+print(paste("AUC at 12 months:", round(auc_val, 3)))
+
+# Step 5: Confusion Matrix (You can choose threshold = 0 by default or use Youden index)
+predicted_class <- ifelse(risk_scores > 0, 1, 0)
+
+# Confusion Matrix
+conf_matrix <- confusionMatrix(factor(predicted_class), factor(true_class), positive = "1")
+print(conf_matrix)
+
+```
+
+## Comparing the models
+
+I compared three modeling approaches to predict 12-month survival outcomes: Random Survival Forest (RSF), LASSO Cox regression, and traditional Random Forest (RF).
+RSF achieved the highest AUC (0.74), indicating superior discrimination, and balanced sensitivity (86.5%) and specificity (47.9%).
+LASSO Cox achieved slightly lower AUC (0.66) but had comparable sensitivity and slightly better accuracy.
+The RF classifier performed poorly (AUC = 0.39), highlighting the limitations of using classification models for censored survival data.
+RSF is optimal for prediction, while LASSO offers interpretability and variable selection.
+
+```{r}
+####################################################
+# Compare RSF vs LASSO Cox: AUC & Confusion Matrix
+# Fixed time point: 12 months
+####################################################
+
+# Step 1: Common dataset for both models
+eval_data <- rsf_data %>%
+  select(survival_months, event, mets_lung, mets_liver, mets_bone, mets_brain,
+         age_group, race, median_income, diagnosis_year) %>%
+  filter(if_all(everything(), ~ !is.na(.))) %>%
+  filter(survival_months > 0)
+
+# Define event status by 12 months
+target_time <- 12
+true_class <- ifelse(eval_data$survival_months <= target_time & eval_data$event == 1, 1, 0)
+
+##############################################
+# Part 1: RSF Model Predictions
+##############################################
+# Predict survival probabilities
+rsf_preds <- predict(rsf_model, data = eval_data, type = "response")
+rsf_surv12 <- rsf_preds$survival[, which.min(abs(rsf_model$unique.death.times - target_time))]
+rsf_pred_class <- ifelse(rsf_surv12 < 0.5, 1, 0)
+
+# AUC
+rsf_roc <- roc(response = true_class, predictor = 1 - rsf_surv12)
+rsf_auc <- auc(rsf_roc)
+print(paste("RSF AUC:", round(rsf_auc, 3)))
+
+# Confusion Matrix
+rsf_cm <- confusionMatrix(factor(rsf_pred_class), factor(true_class), positive = "1")
+print(rsf_cm)
+
+##############################################
+# Part 2: LASSO Cox Model Predictions
+##############################################
+# Create model matrix
+x_lasso <- model.matrix(~ mets_lung + mets_liver + mets_bone + mets_brain +
+                          age_group + race + median_income + diagnosis_year,
+                        data = eval_data)[, -1]
+
+# Linear risk scores from LASSO Cox
+lasso_scores <- predict(cvfit, newx = x_lasso, s = "lambda.min", type = "link")
+lasso_pred_class <- ifelse(lasso_scores > 0, 1, 0)
+
+# AUC
+lasso_roc <- roc(response = true_class, predictor = as.vector(lasso_scores))
+lasso_auc <- auc(lasso_roc)
+print(paste("LASSO Cox AUC:", round(lasso_auc, 3)))
+
+# Confusion Matrix
+lasso_cm <- confusionMatrix(factor(lasso_pred_class), factor(true_class), positive = "1")
+print(lasso_cm)
+
+```

### 2025-04-03 products/manuscript/Manuscript.qmd
part 4 project

@@
 
+### Machine leaning model
+
+I applied a Random Survival Forest (RSF) model to evaluate the impact of clinical and demographic variables on overall survival time.
+RSF is a non-parametric ensemble learning method specifically designed for right-censored survival data.
+It extends the traditional random forest approach by using survival splitting rules and ensemble cumulative hazard estimates.
+The outcome of interest was overall survival time in months, with vital status (0 = alive, 1 = dead) as the event indicator.
+Predictor variables included metastasis status (bone, liver, lung, brain), age group, race, year of diagnosis, and median household income.
+All categorical variables were converted to factors, as required by the randomForestSRC package in R.
+The model was trained using 500 trees, and variable importance was assessed using the impurity measure.
+Model performance was evaluated through time-dependent ROC analysis, confusion matrix, and individualized survival curve prediction.
+The RSF approach enabled flexible modeling of complex interactions and non-linear effects, offering an interpretable and robust alternative to traditional Cox models.
+
 {{< pagebreak >}}
@@
 ## Basic statistical analysis
+
 ### Bivariate Analysis: Metastasis and Survival
@@
 ```
+
 ### Bivariate Analysis: Race and Survival
-Patients identified as  had significantly worse survival outcomes than White patients, with a 37% higher risk of death (HR = 1.37, p < 0.0001).
+
+Patients identified as had significantly worse survival outcomes than White patients, with a 37% higher risk of death (HR = 1.37, p \< 0.0001).
 In contrast, Asian patients had a lower risk of death compared to White patients (HR = 0.86, p = 0.0076), suggesting a potential survival advantage.
@@
 @tbl-summaryhrrace shows the HR for patients with different race.
+
 ```{r}
@@
 ```
+
 ### Bivariate Analysis: Age and Survival
+
 The hazard ratio (HR) estimates for different age groups indicate a general trend of increasing risk with age, though statistical significance varies.
@@
 @tbl-summaryhrage shows the HR for diferrent age groups.
+
 ```{r}
@@
 ```
+
 ### Multivariable analysis
 
-The multivariable Cox proportional hazards model results indicate that metastasis to the bone (HR = 1.30, p < 0.001), brain (HR = 1.72, p < 0.001), liver (HR = 1.66, p < 0.001), and lung (HR = 1.51, p < 0.001) are all significantly associated with poorer survival outcomes.
-Among racial groups, Black patients (HR = 1.39, p < 0.001) had significantly worse survival compared to White patients, while Asian or Pacific Islander and American Indian/Alaska Native groups did not show a statistically significant difference.
+The multivariable Cox proportional hazards model results indicate that metastasis to the bone (HR = 1.30, p \< 0.001), brain (HR = 1.72, p \< 0.001), liver (HR = 1.66, p \< 0.001), and lung (HR = 1.51, p \< 0.001) are all significantly associated with poorer survival outcomes.
+Among racial groups, Black patients (HR = 1.39, p \< 0.001) had significantly worse survival compared to White patients, while Asian or Pacific Islander and American Indian/Alaska Native groups did not show a statistically significant difference.
 
@@
 Year of diagnosis did not show a consistent or significant trend, suggesting no clear improvement or decline in survival based on the year of diagnosis.
-Household income was associated with survival, with patients in the $50,000 - $54,999 (HR = 1.16, p = 0.011) and $65,000 - $69,999 (HR = 1.12, p = 0.030) categories showing significantly higher risk compared to those earning $75,000+.
+Household income was associated with survival, with patients in the \$50,000 - \$54,999 (HR = 1.16, p = 0.011) and \$65,000 - \$69,999 (HR = 1.12, p = 0.030) categories showing significantly higher risk compared to those earning \$75,000+.
 However, other income groups did not show significant differences.
@@
 
-## Full analysis
+### Random Survival Forest
 
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
+A Random Survival Forest (RSF) model was used to evaluate the contribution of clinical and demographic predictors to overall survival.
+The RSF model demonstrated strong performance in predicting survival at 12 months, achieving an AUC of 0.74 (@fig-rsf-auc).
+The confusion matrix revealed a sensitivity of 86.5% and a specificity of 47.9%, with an overall accuracy of 53.8% and balanced accuracy of 67.2%, indicating that the model was effective in identifying patients at high risk of death within the first year.
 
-Example @tbl-resulttable2 shows a summary of a linear model fit.
+Variable importance analysis revealed that age group, median household income, and year of diagnosis were the most influential predictors of survival (@fig-rsf-variable-importance).
+Other important variables included race and metastasis to the bone, lung, liver, and brain.
+The individualized survival prediction for a representative patient showed a steady decline in survival probability over time, reflecting the nature of long-term risk in this cohort (@fig-rsf-ind-sur-pre).
+@fig-rsf-con-matrix.
+Confusion matrix showing prediction results of the Random Survival Forest model at 12 months.
+The model correctly classified 709 survivors and 2,182 non-survivors, with moderate misclassification of 2,376 survivors and 111 non-survivors.
 
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| label: fig-rsf-auc
+#| fig-cap: "Random Survival Forest AUC"
 #| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
+knitr::include_graphics(here("results", "figures", "rsf_auc_12mo.png"))
 ```
 
+```{r}
+#| label: fig-rsf-variable-importance
+#| fig-cap: "Random Survival Forest Variable Importance"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "rsf_variable_importance.png"))
+```
+```{r}
+#| label: fig-rsf-ind-sur-pre
+#| fig-cap: "Random Survival Forest Individualized Survival Prediction"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "rsf_survival_curve_patient1.png"))
+```
+```{r}
+#| label: fig-rsf-con-matrix
+#| fig-cap: "RSF Confusion Matrix (12 Months)"
+#| echo: FALSE
+knitr::include_graphics(here("results", "figures", "rsf_conf_matrix_12mo.png"))
+```
 {{< pagebreak >}}

### 2025-04-18 code/analysis-code/statistical-analysis.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

@@
 library(glmnet)
+library(gt)
+library(dplyr)
+library(here)
+library(webshot2)
+library(fs)
 ```
@@
 saveRDS(cox_results, file = here::here("results", "tables", "mets_HR.rds"))
+
+```
+```{r}
+# Publication Quality table
+# Step 1: Create survival object (if not already created)
+# Replace with your actual survival time and event columns
+# surv_obj <- Surv(time = mydata_filtered$survival_months, event = mydata_filtered$death_flag)
+
+# Step 2: Run Cox regression function with conf.int = TRUE
+run_cox <- function(var) {
+  formula <- as.formula(paste("surv_obj ~", paste0("`", var, "`")))
+  cox_model <- coxph(formula, data = mydata_filtered)
+  result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)
+  return(result)
+}
+
+# Step 3: Run the function for each metastasis site
+cox_bone  <- run_cox("mets_bone")
+cox_brain <- run_cox("mets_brain")
+cox_liver <- run_cox("mets_liver")
+cox_lung  <- run_cox("mets_lung")
+
+# Step 4: Combine and relabel terms
+cox_results <- bind_rows(cox_bone, cox_brain, cox_liver, cox_lung) %>%
+  mutate(term = case_when(
+    term == "mets_boneYes"  ~ "Bone Metastasis",
+    term == "mets_brainYes" ~ "Brain Metastasis",
+    term == "mets_liverYes" ~ "Liver Metastasis",
+    term == "mets_lungYes"  ~ "Lung Metastasis",
+    TRUE ~ term
+  ))
+
+# Step 5: Format for presentation
+cox_formatted <- cox_results %>%
+  mutate(
+    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
+    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
+    significant = ifelse(p.value < 0.05, TRUE, FALSE)
+  ) %>%
+  select(Metastasis = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, significant)
+
+# Step 6: Create GT table with red highlight for significant rows
+# Format for presentation but do not include `significant` column in final output
+cox_formatted <- cox_results %>%
+  mutate(
+    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
+    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
+    highlight = ifelse(p.value < 0.05, TRUE, FALSE)  # internal use only
+  )
+
+# GT table without the highlight column visible
+mets_cox_gt <- cox_formatted %>%
+  select(Metastasis = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, highlight) %>%
+  gt() %>%
+  tab_header(
+    title = md("**Table 5: Association Between Metastasis and Overall Survival**")
+  ) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  tab_style(
+    style = cell_text(color = "red"),
+    locations = cells_body(
+      columns = c(`Hazard Ratio (95% CI)`),
+      rows = highlight == TRUE
+    )
+  ) %>%
+  opt_row_striping() %>%
+  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
+  tab_options(
+    table.font.size = 11,
+    heading.title.font.size = 13,
+    data_row.padding = px(4),
+    table.align = "center",
+    table_body.hlines.color = "gray90",
+    table.border.top.color = "gray80",
+    table.border.bottom.color = "gray80"
+  ) %>%
+  tab_source_note(md("*Note:* Statistically significant hazard ratios (p < 0.05) are shown in red.")) %>%
+  cols_hide(columns = c(highlight))  # 👈 Hides the internal logic column
+
+# Step 7: Save outputs
+saveRDS(cox_results, here("results", "tables", "mets_HR.rds"))
+saveRDS(mets_cox_gt, here("results", "tables", "mets_HR_gt.rds"))
+
+dir_create(here("results", "figures"))
+
+gtsave(mets_cox_gt,
+       filename = here("results", "figures", "mets_HR_table.png"),
+       vwidth = 1000, vheight = 600, zoom = 2)
+#View table here
+mets_cox_gt
 
@@
 saveRDS(cox_race_results, file = here::here("results", "tables", "race_HR.rds"))
+
+```
+```{r}
+#Publication-ready Table 
+# Step 1: Recode the race variable (if not already done)
+mydata_filtered <- mydata_filtered %>%
+  mutate(race = factor(race, levels = c(
+    "White", "Black", "Asian or Pacific Islander", "American Indian/Alaska Native"
+  )))
+
+# Step 2: Define the Cox model function (includes CI)
+run_cox_race <- function() {
+  formula <- as.formula("surv_obj ~ race")
+  cox_model <- coxph(formula, data = mydata_filtered)
+  result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)  # ✅ Includes CI
+  return(result)
+}
+
+# Step 3: Run the model
+cox_race_results <- run_cox_race()
+
+# Step 4: Rename comparisons
+cox_race_results <- cox_race_results %>%
+  mutate(term = case_when(
+    term == "raceBlack" ~ "Black vs. White",
+    term == "raceAsian or Pacific Islander" ~ "Asian vs. White",
+    term == "raceAmerican Indian/Alaska Native" ~ "American Indian vs. White",
+    TRUE ~ term
+  ))
+
+# Step 5: Format HR (95% CI) and significance
+cox_race_formatted <- cox_race_results %>%
+  mutate(
+    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
+    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
+    highlight = ifelse(p.value < 0.05, TRUE, FALSE)
+  )
+
+# Step 6: Create GT table with red highlights
+race_cox_gt <- cox_race_formatted %>%
+  select(Comparison = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, highlight) %>%
+  gt() %>%
+  tab_header(
+    title = md("**Table 6: Association Between Race and Overall Survival**")
+  ) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  tab_style(
+    style = cell_text(color = "red"),
+    locations = cells_body(
+      columns = c(`Hazard Ratio (95% CI)`),
+      rows = highlight == TRUE
+    )
+  ) %>%
+  opt_row_striping() %>%
+  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
+  tab_options(
+    table.font.size = 11,
+    heading.title.font.size = 13,
+    data_row.padding = px(4),
+    table.align = "center",
+    table_body.hlines.color = "gray90",
+    table.border.top.color = "gray80",
+    table.border.bottom.color = "gray80"
+  ) %>%
+  tab_source_note(md("*Note:* Statistically significant hazard ratios (p < 0.05) are shown in red.")) %>%
+  cols_hide(columns = c(highlight))
+
+# Step 7: Show the table in QMD output
+race_cox_gt
+
+# Step 8: Save outputs
+saveRDS(cox_race_formatted, here("results", "tables", "race_HR.rds"))
+saveRDS(race_cox_gt, here("results", "tables", "race_HR_gt.rds"))
+
+dir_create(here("results", "figures"))
+gtsave(race_cox_gt,
+       filename = here("results", "figures", "race_HR_table.png"),
+       vwidth = 1000, vheight = 600, zoom = 2)
+
+# Show table in HTML/preview
+race_cox_gt
 
@@
     str_detect(term, "age_group80-84 years") ~ "80-84",
-    str_detect(term, "age_group85+ years") ~ "85+",
+    str_detect(term, "age_group85\\+ years") ~ "85+",
     TRUE ~ term  # Keep other terms unchanged
@@
 
+```
+```{r}
+#Publication-ready table
+
+# Step 1: Re-run the Cox model with confidence intervals
+run_cox_age <- function() {
+  formula <- as.formula("surv_obj ~ age_group")
+  cox_model <- coxph(formula, data = mydata_filtered)
+  result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE)  # <-- CI INCLUDED
+  return(result)
+}
+
+cox_age_results <- run_cox_age()
+
+# Step 2: Relabel age group comparisons (optional cleaning)
+cox_age_results <- cox_age_results %>%
+  mutate(term = str_replace_all(term, "`", "")) %>%
+  mutate(term = case_when(
+    str_detect(term, "age_group25-29 years") ~ "25–29",
+    str_detect(term, "age_group30-34 years") ~ "30–34",
+    str_detect(term, "age_group35-39 years") ~ "35–39",
+    str_detect(term, "age_group40-44 years") ~ "40–44",
+    str_detect(term, "age_group45-49 years") ~ "45–49",
+    str_detect(term, "age_group50-54 years") ~ "50–54",
+    str_detect(term, "age_group55-59 years") ~ "55–59",
+    str_detect(term, "age_group60-64 years") ~ "60–64",
+    str_detect(term, "age_group65-69 years") ~ "65–69",
+    str_detect(term, "age_group70-74 years") ~ "70–74",
+    str_detect(term, "age_group75-79 years") ~ "75–79",
+    str_detect(term, "age_group80-84 years") ~ "80–84",
+    str_detect(term, "age_group85\\+ years") ~ "85+",
+    TRUE ~ term
+  ))
+
+# Step 3: Format table with HR, CI, p-value, significance flag
+cox_age_formatted <- cox_age_results %>%
+  mutate(
+    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), "–", round(conf.high, 2), ")"),
+    p_value = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
+    highlight = ifelse(p.value < 0.05, TRUE, FALSE)
+  )
+
+# Step 4: Create GT table
+age_cox_gt <- cox_age_formatted %>%
+  select(`Age Group (vs. 20–24)` = term, `Hazard Ratio (95% CI)` = HR_CI, `p-value` = p_value, highlight) %>%
+  gt() %>%
+  tab_header(
+    title = md("**Table 6: Association Between Age Group and Overall Survival**")
+  ) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  tab_style(
+    style = cell_text(color = "red"),
+    locations = cells_body(
+      columns = c(`Hazard Ratio (95% CI)`),
+      rows = highlight == TRUE
+    )
+  ) %>%
+  opt_row_striping() %>%
+  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
+  tab_options(
+    table.font.size = 11,
+    heading.title.font.size = 13,
+    data_row.padding = px(4),
+    table.align = "center",
+    table_body.hlines.color = "gray90",
+    table.border.top.color = "gray80",
+    table.border.bottom.color = "gray80"
+  ) %>%
+  tab_source_note(md("*Note:* Statistically significant hazard ratios (p < 0.05) are shown in red.")) %>%
+  cols_hide(columns = c(highlight))
+
+# Step 5: Save GT object and PNG figure
+saveRDS(age_cox_gt, here("results", "tables", "age_HR_gt.rds"))
+
+dir_create(here("results", "figures"))
+
+gtsave(age_cox_gt,
+       filename = here("results", "figures", "age_HR_table.png"),
+       vwidth = 1000, vheight = 600, zoom = 2)
+
+# Optional: display in output
+age_cox_gt
 ```
@@
 
-# Optional: Save selected variables
-write.csv(lasso_selected, "results/tables/lasso_selected_variables.csv", row.names = FALSE)
-
+saveRDS(lasso_selected, file = here::here("results", "tables", "lasso_selected.rds"))
+saveRDS(cvfit, file = here::here("results", "tables", "cvfit_lasso.rds"))
 ```

### 2025-04-18 code/eda-code/edamz.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

@@
 library(ggplot2)
+library(gt)
+library(webshot2)
+library(fs)
+library(dplyr)
 ```
@@
 ```{r}
+# Publication ready-table
+# Get skim summary
+summary_df <- skim(mydata)
+
+# Filter only for survival_months (numeric type)
+surv_summary <- summary_df %>%
+  filter(skim_variable == "survival_months") %>%
+  select(
+    Variable = skim_variable,
+    Completeness = complete_rate,
+    Mean = numeric.mean,
+    SD = numeric.sd,
+    Min = numeric.p0,
+    Q1 = numeric.p25,
+    Median = numeric.p50,
+    Q3 = numeric.p75,
+    Max = numeric.p100
+  )
+
+# Format using GT
+surv_summary_gt <- surv_summary %>%
+  gt() %>%
+  tab_header(title = md("**Table 2: Summary of Survival Months**")) %>%
+  fmt_number(columns = where(is.numeric), decimals = 2) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
+  tab_options(
+    table.font.size = 11,
+    heading.title.font.size = 13,
+    table_body.hlines.color = "gray90",
+    table.border.top.color = "gray80",
+    table.border.bottom.color = "gray80",
+    row.striping.include_table_body = TRUE,
+    data_row.padding = px(4),
+    table.align = "center"
+  ) %>%
+  opt_row_striping() %>%
+  tab_source_note(md("*Note:* Summary based on numeric variable `survival_months`."))
+
+# Display the table 
+surv_summary_gt
+
+# Save GT object and PNG
+saveRDS(surv_summary_gt, here("results", "edatables", "edasurvivalmonths_table_gt.rds"))
+
+# Ensure image folder exists
+dir_create(here("results", "edafigures"))
+
+# Save as PNG
+gtsave(surv_summary_gt,
+       filename = here("results", "edafigures", "edasurvivalmonths_table.png"),
+       vwidth = 1000, vheight = 600, zoom = 2)
+```
+
+```{r}
 # Create a structured summary table
@@
 saveRDS(var_summary, file = summarytable_file)
+```
+```{r}
+# Create publication-ready table
+var_summary_gt <- var_summary %>%
+  gt() %>%
+  tab_header(
+    title = md("**Table 1: Variable Summary**")
+  ) %>%
+  cols_label(
+    Variable = "Variable Name",
+    Type = "Data Type",
+    Unique_Values = "Unique Values",
+    Missing_Values = "Missing Values"
+  ) %>%
+  fmt_number(
+    columns = vars(Unique_Values, Missing_Values),
+    decimals = 0
+  ) %>%
+  tab_options(
+    table.font.size = "small",
+    heading.title.font.size = 12,
+    data_row.padding = px(4),
+    table.align = "center"
+  )
+
+# Display the table
+var_summary_gt
+
+# Save the GT table object to RDS
+summarytable_file_gt <- here("results", "edatables", "edasummarytable2_gt.rds")
+# Save table as GT object (already done)
+saveRDS(var_summary_gt, file = here("results", "edatables", "edasummarytable2_gt.rds"))
+
+# ALSO save as high-resolution PNG
+gtsave(var_summary_gt,
+       filename = here("results", "edafigures", "summarytable.png"),
+       vwidth = 1200, vheight = 800, zoom = 2)
+
 ```
@@
 ```
+```{r}
+# Generate summary Publication ready table for age_group
+p1 <- mydata %>%
+  group_by(age_group) %>%
+  summarise(Count = n(), .groups = "drop")
+
+# Style it using gt
+age_summary_gt <- p1 %>%
+  gt() %>%
+  tab_header(
+    title = md("**Table 3: Distribution of Age Groups**")
+  ) %>%
+  cols_label(
+    age_group = "Age Group",
+    Count = "Number of Patients"
+  ) %>%
+  fmt_number(columns = Count, decimals = 0) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  opt_row_striping() %>%
+  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
+  tab_options(
+    table.font.size = 11,
+    heading.title.font.size = 13,
+    data_row.padding = px(4),
+    table.align = "center",
+    table_body.hlines.color = "gray90",
+    table.border.top.color = "gray80",
+    table.border.bottom.color = "gray80"
+  ) %>%
+  tab_source_note(md("*Note:* This table summarizes the distribution of patients by age group."))
+
+# Show in EDA HTML output
+age_summary_gt
+
+# Save raw table
+saveRDS(p1, here("results", "edatables", "edaagesummary.rds"))
+
+# Save styled table
+saveRDS(age_summary_gt, here("results", "edatables", "edaagesummary_gt.rds"))
+
+# Ensure folder exists and save PNG
+dir_create(here("results", "edafigures"))
+
+gtsave(age_summary_gt,
+       filename = here("results", "edafigures", "edaagesummary.png"),
+       vwidth = 800, vheight = 500, zoom = 2)
+```
+
 Now lets look at the bar plot for age distribution
@@
 ```
+```{r}
+#Publication ready table 
+# Generate clean summary of SEER Combined Mets at Diagnosis
+# Clean and relabel the metastasis site names
+p7 <- mydata %>%
+  summarise(
+    Bone_Mets  = sum(mets_bone == "Yes", na.rm = TRUE),
+    Brain_Mets = sum(mets_brain == "Yes", na.rm = TRUE),
+    Liver_Mets = sum(mets_liver == "Yes", na.rm = TRUE),
+    Lung_Mets  = sum(mets_lung == "Yes", na.rm = TRUE)
+  ) %>%
+  pivot_longer(cols = everything(), names_to = "Metastasis_Site", values_to = "Count") %>%
+  mutate(Metastasis_Site = case_when(
+    Metastasis_Site == "Bone_Mets"  ~ "Bone Metastasis",
+    Metastasis_Site == "Brain_Mets" ~ "Brain Metastasis",
+    Metastasis_Site == "Liver_Mets" ~ "Liver Metastasis",
+    Metastasis_Site == "Lung_Mets"  ~ "Lung Metastasis"
+  ))
+
+# Format the GT table
+mets_summary_gt <- p7 %>%
+  gt() %>%
+  tab_header(title = md("**Table 4: Metastasis at Diagnosis by Site**")) %>%
+  cols_label(
+    Metastasis_Site = "Site",
+    Count = "Number of Patients"
+  ) %>%
+  fmt_number(columns = Count, decimals = 0) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels(everything())
+  ) %>%
+  opt_table_font(font = list(google_font("Roboto"), default_fonts())) %>%
+  opt_row_striping() %>%
+  tab_options(
+    table.font.size = 11,
+    heading.title.font.size = 13,
+    data_row.padding = px(4),
+    table.align = "center",
+    table_body.hlines.color = "gray90",
+    table.border.top.color = "gray80",
+    table.border.bottom.color = "gray80"
+  ) %>%
+  tab_source_note(md("*Note:* This table summarizes the number of patients with bone, brain, liver, or lung metastases at diagnosis."))
+
+# Show in EDA output
+mets_summary_gt
+
+# Save RDS versions
+saveRDS(p7, here("results", "edatables", "edametsatsummary.rds"))
+saveRDS(mets_summary_gt, here("results", "edatables", "edametsatsummary_gt.rds"))
+
+# Save PNG version for manuscript
+dir_create(here("results", "edafigures"))
+
+gtsave(mets_summary_gt,
+       filename = here("results", "edafigures", "edametsatsummary.png"),
+       vwidth = 900, vheight = 600, zoom = 2)
+```
 

### 2025-04-18 code/processing-code/processingcodemz.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

@@
 ## Renaming all the variables for the ease of analysis
+
 ```{r}

### 2025-04-18 products/manuscript/Manuscript.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

@@
 library(knitr)
+library(gt)
+library(webshot2)
 ```
@@
 
-# Summary/Abstract
+# Abstract
 
-*Write a summary of your project.*
+**Background:** Endometrial cancer is the most common gynecological malignancy in the United States.
+While most cases are diagnosed early, stage IV disease presents significant prognostic challenges, particularly when distant metastases are involved.
+
+**Objective:** This study aimed to evaluate clinicopathological characteristics and survival outcomes among patients with stage IV endometrial cancer, focusing on differences based on metastatic site (bone, brain, liver, lung).
+
+**Methods:** We conducted a retrospective cohort analysis using SEER data from 2010–2015.
+Kaplan-Meier and Cox proportional hazards models were used to assess survival differences, and a Random Survival Forest (RSF) model was implemented to evaluate variable importance and predict survival probabilities.
+
+**Results:** Among 5,708 patients with stage IV endometrial cancer, lung metastases were most common (29.4%), followed by liver (14.8%), bone (9.7%), and brain (2.8%).
+Median survival was 12 months.
+In multivariable analysis, all metastasis types were significantly associated with poorer survival, especially brain (HR = 1.72) and liver (HR = 1.66).
+Older age and Black race were also linked to worse outcomes.
+The RSF model achieved an AUC of 0.74, with age group, income, and metastatic site identified as the most important survival predictors.
+
+**Conclusions:** Metastatic site significantly influences survival in stage IV endometrial cancer.
+Tailored interventions and surveillance strategies based on metastasis patterns and patient demographics may help optimize outcomes.
 
@@
 #| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS(here("results","edatables","edasummarytable2.rds"))
-knitr::kable(resulttable)
+#| echo: false
+knitr::include_graphics(here::here("results", "edafigures", "summarytable.png"))
 ```
@@
 #| label: tbl-summarytablesurvival
-#| tbl-cap: "Summary of Survival."
-#| echo: FALSE
-resulttable=readRDS(here("results","edatables","edasurvivalmonthssummary.rds"))
-knitr::kable(resulttable)
+#| tbl-cap: "Summary of survival months in the study population."
+#| echo: false
+knitr::include_graphics(here::here("results", "edafigures", "edasurvivalmonths_table.png"))
 ```
@@
 #| label: tbl-summaryage
-#| tbl-cap: "Summary of age."
-#| echo: FALSE
-resulttable=readRDS(here("results","edatables","edaagesummary.rds"))
-knitr::kable(resulttable)
+#| tbl-cap: "Summary of age group distribution."
+#| echo: false
+knitr::include_graphics(here::here("results", "edafigures", "edaagesummary.png"))
 ```
@@
 #| echo: FALSE
-resulttable=readRDS(here("results","edatables","edametsatsummary.rds"))
-knitr::kable(resulttable)
+knitr::include_graphics(here::here("results", "edafigures", "edametsatsummary.png"))
 ```
@@
 #| echo: FALSE
-knitr::include_graphics(here("results","edafigures","edametsatdxdistribution.png"))
+knitr::include_graphics(here::here("results", "edafigures", "edametsatdxdistribution.png"))
 ```
@@
 #| echo: FALSE
-resulttable=readRDS(here("results", "tables", "mets_HR.rds"))
-knitr::kable(resulttable)
+knitr::include_graphics(here::here("results", "figures", "mets_HR_table.png"))
 ```
@@
 #| echo: FALSE
-resulttable=readRDS(here("results", "tables", "race_HR.rds"))
-knitr::kable(resulttable)
+knitr::include_graphics(here::here("results", "figures", "race_HR_table.png"))
 ```
@@
 #| echo: FALSE
-resulttable=readRDS(here("results", "tables", "age_HR.rds"))
-knitr::kable(resulttable)
+knitr::include_graphics(here::here("results", "figures", "age_HR_table.png"))
 ```
@@
 The individualized survival prediction for a representative patient showed a steady decline in survival probability over time, reflecting the nature of long-term risk in this cohort (@fig-rsf-ind-sur-pre).
-@fig-rsf-con-matrix.
+(@fig-rsf-con-matrix).
 Confusion matrix showing prediction results of the Random Survival Forest model at 12 months.
@@
 ```
+
 ```{r}
@@
 ```
+
 {{< pagebreak >}}
@@
 
-*Summarize what you did, what you found and what it means.*
+In this large SEER-based analysis of patients with stage IV endometrial cancer, we found that survival outcomes varied significantly based on the site of metastasis.
+Lung metastasis was the most common at diagnosis, consistent with the known hematogenous spread of endometrial tumors to the pulmonary system.
+However, brain and liver metastases were associated with the poorest prognoses, showing significantly elevated hazard ratios for mortality, even after adjusting for age, race, income, and year of diagnosis.
+This suggests that not all metastatic patterns carry equal prognostic weight, and involvement of certain visceral organs may reflect more aggressive tumor biology or greater resistance to conventional treatment modalities.
+These findings align with previous literature highlighting the aggressive nature and poor survival associated with visceral metastases, particularly in the central nervous system and liver, where treatment options are often limited and complications more severe [@Mao2020; @el2023patterns].
+
+Our study also highlights important demographic disparities that influence survival.
+Black patients had significantly higher mortality risk compared to White patients, a pattern consistent with prior studies on racial inequities in cancer detection, treatment access, and outcomes [@Monk2022; @mccarthy2022racial].
+These disparities may reflect differences in healthcare access, socioeconomic factors, and biological variations, and underscore the need for tailored interventions to address inequities in care delivery.
+The impact of age was particularly striking—patients aged ≥85 years had more than 12 times the mortality risk of younger patients.
+This finding supports the well-established role of age as a dominant prognostic factor in gynecologic malignancies, not only due to physiological frailty and comorbidities but also possibly due to undertreatment in older populations [@MORICE20161094; @smith2023impact].
+As the population ages, understanding and addressing these age-related disparities will be increasingly important for improving outcomes in this vulnerable group.
+
+The application of a Random Survival Forest (RSF) model in our analysis provided additional insights beyond those offered by traditional Cox regression.
+By capturing complex, non-linear relationships and variable interactions, the RSF approach identified the most influential predictors of survival—including age group, income, and metastatic site.
+This machine learning method yielded an AUC of 0.74, indicating good predictive performance and offering a robust complement to the conventional modeling framework.
+Notably, RSF enabled individualized survival prediction, which may have potential utility in clinical decision-making and patient counseling [@ishwaran2008random; @wager2017estimation].
+Integrating such tools into routine care could enhance risk stratification, support shared decision-making, and contribute to the development of precision oncology strategies.
 
@@
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
+One of the study’s major strengths is the use of the SEER registry, a high-quality, population-based dataset capturing data across a wide U.S. demographic.
+This enhances the generalizability of our findings and allows for robust subgroup analyses across metastasis types and sociodemographic strata.
+[@howlader2021seer] Moreover, the combination of classical survival analysis and machine learning techniques provides both interpretability and predictive power.
+
+However, our study has several limitations.
+The SEER database lacks treatment details such as chemotherapy, hormonal therapy, or immunotherapy, which may influence survival outcomes.
+Also, performance status and comorbidities—key clinical predictors—are not captured in SEER.
+The data only reflect metastasis at diagnosis, not accounting for progression or treatment response.
+Lastly, despite adjusting for income and race, residual confounding due to unmeasured variables is possible.
 
@@
 
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+Our findings suggest that survival in stage IV endometrial cancer is heavily influenced by the site of metastasis, with brain and liver involvement predicting worse outcomes.
+The disparities across age and race underscore the need for tailored care and equitable access to treatment.
+Incorporating advanced analytics such as RSF provides a deeper understanding of survival dynamics and may inform future personalized prognostic tools.
+Future research should aim to integrate molecular, treatment, and longitudinal clinical data to refine prognostic models.
+These findings could also guide clinicians in risk stratification, resource allocation, and decision-making, especially for high-risk subgroups.
 

### 2025-04-18 products/manuscript/supplement/Supplementary-Material.qmd
Part 5 Manuscript Dr. @andreahandel addressed comments based on issue 4

---
title: "Supplement to Example Manuscript Template for a Data Analysis Project"
format:
  pdf:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Overview

This supplementary material provides additional details, analyses, and outputs supporting the findings presented in the main manuscript. It includes extended methodological explanations, reproducibility instructions, and results from exploratory data analysis. Key supplementary tables and figures include variable summaries, survival modeling outputs, and machine learning performance metrics. Detailed results from Cox regression, Random Forest, and LASSO Cox models are presented along with confusion matrices, AUC values, and variable importance rankings. The structure and content of the project directory are outlined to facilitate full reproducibility. Together, these materials offer a comprehensive view of the analytical workflow and support the robustness of the study’s conclusions.

# Code and file information

This project is organized in a modular and reproducible structure, separating raw data, processing scripts, analysis outputs, and final products. Below is a guide to the directory structure and instructions to reproduce all results.

### Directory Structure Overview

ZUBER-MADA-project/
│
├── data/
│   ├── raw-data/           # Contains the original input datasets (raw files)
│   └── processed-data/     # Contains the cleaned and transformed datasets
│
├── code/
│   ├── processing-code/    # Data cleaning scripts
│   ├── eda-code/           # Exploratory data analysis scripts
│   └── analysis-code/      # Statistical modeling and prediction
│
├── results/
│   ├── edatables/          # Exploratory summary tables
│   ├── edafigures/         # Exploratory plots
│   ├── tables/             # Model results tables
│   ├── figures/            # Model-related figures
│   └── output/             # Rendered final outputs (PDF/HTML)
│
├── products/
│   ├── manuscript/         # Manuscript and supplementary files
│   ├── poster/             # Poster material (if applicable)
│   └── slides/             # Presentation slides (if applicable)
│
├── assets/                 # Bibliography and citation style files
└── README.md               # Project overview and general usage instructions

{{< pagebreak >}}

# Additional Method Details

Often, the main manuscript only allows for an overview description of the methods. Use the supplement to describe all your methods, models and approaches in a lot of detail. Reference specific parts of your code as needed.

{{< pagebreak >}}

# Additional results

Show additional results here. Those can be some useful exploratory/descriptive figures or tables, or results from additional analyses that didn't make it into the main text.

# Additional result

## Exploratory Analysis:Race Distribution
As part of the descriptive exploration of the dataset, we examined the distribution of patients by race. The majority of patients in the cohort identified as White, followed by Black, Asian or Pacific Islander, American Indian/Alaska Native, and a small proportion with unknown race information. @tbl-race-summary. Frequency distribution of patients across racial categories.

```{r}
#| label: tbl-race-summary
#| tbl-cap: "Distribution of patients by race in the analytic cohort."
#| echo: FALSE
p3 <- readRDS(here::here("results", "edatables", "edaracesummary.rds"))
knitr::kable(p3)

```
## Exploratory Summary: Year of Diagnosis

We also summarized the distribution of patients by year of cancer diagnosis to understand the temporal spread of the study cohort. This information helps contextualize potential changes in treatment patterns, diagnostic advancements, or survival outcomes over time. @tbl-year-summary. Frequency distribution of patients by year of diagnosis.
```{r}
#| label: tbl-year-summary
#| tbl-cap: "Distribution of patients by year of diagnosis."
#| echo: FALSE
p5 <- readRDS(here::here("results", "edatables", "edayearsummary.rds"))
knitr::kable(p5)

```
@fig-year-diagnosis. Bar plot displaying the distribution of patients by year of diagnosis. The figure shows a relatively even spread across the years, with some fluctuations that may reflect data entry trends or cohort recruitment variability.
```{r}
#| label: fig-year-diagnosis
#| fig-cap: "Bar plot showing the distribution of patients by year of diagnosis."
#| echo: FALSE
knitr::include_graphics(here::here("results", "edafigures", "edayear.png"))

```

## Exploratory Summary: Cause-Specific Death Classification
We summarized the distribution of outcomes according to the SEER cause-specific death classification. The majority of patients in the dataset (77.5%) were classified as having died due to the primary cancer diagnosis, while 22.5% were either alive or had died from other causes. This outcome classification was used as the primary event definition in survival models throughout the analysis.

@tbl-cause-death. Distribution of patients by cause-specific death classification.

```{r}
#| label: tbl-cause-death
#| tbl-cap: "Cause-specific death classification among patients in the cohort."
#| echo: FALSE
p9 <- readRDS(here::here("results", "edatables", "edacausedeathsummary.rds"))
knitr::kable(p9)

```

## Univariate Analysis

To evaluate whether survival outcomes varied over time, we conducted a bivariate Cox proportional hazards analysis using year of diagnosis as the predictor. The year 2010 was used as the reference category. Hazard ratios (HRs) for each subsequent year (2011 to 2015) were calculated to assess whether patients diagnosed in later years had different risks of death compared to those diagnosed in 2010. As shown in @tbl-year-diagnosis, all HRs were close to 1, and none of the comparisons reached statistical significance (all p \> 0.05), indicating that there was no significant year-to-year variation in survival over the study period. These findings suggest a relative consistency in survival outcomes for patients diagnosed between 2010 and 2015 within this dataset.

```{r}
#| label: tbl-year-diagnosis
#| tbl-cap: "Hazard ratios from Cox regression evaluating year of diagnosis and survival."
#| echo: FALSE
cox_year_results <- readRDS(here::here("results", "tables", "year_HR.rds"))
knitr::kable(cox_year_results, digits = 3)
```

## Random Forest Model: Predicting Cause-Specific Death

We implemented a random forest classifier to predict whether a patient’s death was attributable to cancer using SEER registry variables. The dataset was preprocessed with downsampling to address class imbalance. We used 5-fold cross-validation to tune hyperparameters and selected the best model based on ROC AUC. The model was trained on 80% of the data and evaluated on a held-out 20% test set.

The top predictors, based on variable importance, included lung and liver metastases, age group, and year of diagnosis.

@fig-vip-rf .Variable importance plot from the final tuned random forest model. The top 10 predictors are displayed, with mets_lung and mets_liver being the most influential variables for predicting cause-specific death.

```{r}
#| label: fig-vip-rf
#| fig-cap: "Top 10 most important variables from the random forest model predicting cause-specific death."
#| echo: FALSE
knitr::include_graphics(here::here("results", "figures", "vip_plot_rf_balanced.png"))
```

@tbl-rf-cm. Confusion matrix summarizing the random forest model’s classification performance on the test set. The model shows better performance in identifying cancer-attributable deaths than non-cancer deaths, although overall accuracy was limited.
```{r}
#| label: tbl-rf-cm
#| tbl-cap: "Confusion matrix from the random forest model on the test set."
#| echo: FALSE
rf_cm <- readRDS(here::here("results", "tables", "random_forest_balanced_confusion_matrix.rds"))
rf_cm
```
@tbl-rf-auc. Area under the receiver operating characteristic curve (ROC AUC) from the random forest model. The model achieved an AUC of approximately 0.39, indicating limited discriminative performance, possibly due to overlapping features and class imbalance.
```{r}
#| label: tbl-rf-auc
#| tbl-cap: "ROC AUC score of the final random forest model."
#| echo: FALSE
rf_auc <- readRDS(here::here("results", "tables", "random_forest_balanced_auc.rds"))
rf_auc
```

##  LASSO Cox Regression for Variable Selection.
To identify the most influential predictors of cancer-specific survival, we applied LASSO (Least Absolute Shrinkage and Selection Operator) penalized Cox regression using the glmnet package. This approach helps in variable selection by shrinking less informative coefficients to zero while retaining strong predictors. We performed 10-fold cross-validation to select the optimal penalty parameter (λ), minimizing the partial likelihood deviance.

@tbl-lasso-vars. Selected variables from the LASSO model using the optimal penalty parameter (λ.min). These are the predictors that remained in the final model with non-zero coefficients.

```{r}
#| label: tbl-lasso-vars
#| tbl-cap: "Non-zero coefficients from the LASSO Cox regression model."
#| echo: FALSE
lasso_selected <- readRDS(here::here("results", "tables", "lasso_selected.rds"))
lasso_selected %>% knitr::kable(digits = 4)
```

## LASSO Cox Model Performance at 12 Months
To assess the discriminatory ability of the LASSO Cox regression model, we evaluated model performance at 12 months post-diagnosis. The linear predictor (risk score) was used to generate a receiver operating characteristic (ROC) curve, and area under the curve (AUC) was calculated. Additionally, we computed a confusion matrix using a default threshold of 0 for classification.

@tbl-lasso-auc. Model discrimination at 12 months, measured by area under the ROC curve (AUC). The LASSO Cox model achieved a moderate discriminative ability with an AUC of 0.664.
```{r}
#| label: tbl-lasso-auc
#| tbl-cap: "AUC for the LASSO Cox model at 12-month survival prediction."
#| echo: FALSE
paste("AUC at 12 months:", round(auc_val, 3))

```

@tbl-lasso-cm. Confusion matrix at 12 months using a classification threshold of 0. The model achieved high sensitivity (87.8%) but low specificity (25.7%), indicating it was better at correctly identifying deaths than survivals at this time point.
```{r}
#| label: tbl-lasso-cm
#| tbl-cap: "Confusion matrix for LASSO Cox model predicting 12-month mortality."
#| echo: FALSE
conf_matrix

```

{{< pagebreak >}}

# Discussion

Any additional discussion regarding the supplementary material/findings.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

{{< pagebreak >}}

# References

### 2025-04-18 products/manuscript/supplement/Supplementary-Material.qmd
Part 5 V2

@@
 format:
-  pdf:
+  docx:
     toc: false
@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
 ---
@@
 library(knitr)
+library(tidyverse)
+
 ```
@@
 
-### Directory Structure Overview
+# Directory Structure Overview
 
+### 📁 Directory Structure Overview
+
+```
 ZUBER-MADA-project/
-│
 ├── data/
-│   ├── raw-data/           # Contains the original input datasets (raw files)
-│   └── processed-data/     # Contains the cleaned and transformed datasets
+│   ├── raw-data/           # Original input datasets (raw files)
+│   └── processed-data/     # Cleaned and transformed datasets
 │
@@
 └── README.md               # Project overview and general usage instructions
+```
 
@@
 ```
-
-## LASSO Cox Model Performance at 12 Months
-To assess the discriminatory ability of the LASSO Cox regression model, we evaluated model performance at 12 months post-diagnosis.
-The linear predictor (risk score) was used to generate a receiver operating characteristic (ROC) curve, and area under the curve (AUC) was calculated.
-Additionally, we computed a confusion matrix using a default threshold of 0 for classification.
-
-@tbl-lasso-auc. Model discrimination at 12 months, measured by area under the ROC curve (AUC). The LASSO Cox model achieved a moderate discriminative ability with an AUC of 0.664.
-```{r}
-#| label: tbl-lasso-auc
-#| tbl-cap: "AUC for the LASSO Cox model at 12-month survival prediction."
-#| echo: FALSE
-paste("AUC at 12 months:", round(auc_val, 3))
-
-```
-
-@tbl-lasso-cm.
-Confusion matrix at 12 months using a classification threshold of 0.
-The model achieved high sensitivity (87.8%) but low specificity (25.7%), indicating it was better at correctly identifying deaths than survivals at this time point.
-```{r}
-#| label: tbl-lasso-cm
-#| tbl-cap: "Confusion matrix for LASSO Cox model predicting 12-month mortality."
-#| echo: FALSE
-conf_matrix
-
-```
-
-{{< pagebreak >}}
-
-# Discussion
-
-Any additional discussion regarding the supplementary material/findings.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-{{< pagebreak >}}
-
-# References

### 2025-04-18 products/manuscript/supplement/Supplementary-Material.tex
Part 5 V2

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Supplement to Example Manuscript Template for a Data Analysis Project},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Supplement to Example Manuscript Template for a Data Analysis
Project}
\author{}
\date{}

\begin{document}
\maketitle

\section{Overview}\label{overview}

This supplementary material provides additional details, analyses, and
outputs supporting the findings presented in the main manuscript. It
includes extended methodological explanations, reproducibility
instructions, and results from exploratory data analysis. Key
supplementary tables and figures include variable summaries, survival
modeling outputs, and machine learning performance metrics. Detailed
results from Cox regression, Random Forest, and LASSO Cox models are
presented along with confusion matrices, AUC values, and variable
importance rankings. The structure and content of the project directory
are outlined to facilitate full reproducibility. Together, these
materials offer a comprehensive view of the analytical workflow and
support the robustness of the study's conclusions.

\section{Code and file information}\label{code-and-file-information}

This project is organized in a modular and reproducible structure,
separating raw data, processing scripts, analysis outputs, and final
products. Below is a guide to the directory structure and instructions
to reproduce all results.

\subsubsection{Directory Structure
Overview}\label{directory-structure-overview}

ZUBER-MADA-project/ │ ├── data/ │ ├── raw-data/ \# Contains the original
input datasets (raw files) │ └── processed-data/ \# Contains the cleaned
and transformed datasets │ ├── code/ │ ├── processing-code/ \# Data
cleaning scripts │ ├── eda-code/ \# Exploratory data analysis scripts │
└── analysis-code/ \# Statistical modeling and prediction │ ├── results/
│ ├── edatables/ \# Exploratory summary tables │ ├── edafigures/ \#
Exploratory plots │ ├── tables/ \# Model results tables │ ├── figures/
\# Model-related figures │ └── output/ \# Rendered final outputs
(PDF/HTML) │ ├── products/ │ ├── manuscript/ \# Manuscript and
supplementary files │ ├── poster/ \# Poster material (if applicable) │
└── slides/ \# Presentation slides (if applicable) │ ├── assets/ \#
Bibliography and citation style files └── README.md \# Project overview
and general usage instructions

\newpage{}

\section{Additional Method Details}\label{additional-method-details}

Often, the main manuscript only allows for an overview description of
the methods. Use the supplement to describe all your methods, models and
approaches in a lot of detail. Reference specific parts of your code as
needed.

\newpage{}

\section{Additional results}\label{additional-results}

Show additional results here. Those can be some useful
exploratory/descriptive figures or tables, or results from additional
analyses that didn't make it into the main text.

\section{Additional result}\label{additional-result}

\subsection{Exploratory Analysis:Race
Distribution}\label{exploratory-analysisrace-distribution}

As part of the descriptive exploration of the dataset, we examined the
distribution of patients by race. The majority of patients in the cohort
identified as White, followed by Black, Asian or Pacific Islander,
American Indian/Alaska Native, and a small proportion with unknown race
information. Table~\ref{tbl-race-summary}. Frequency distribution of
patients across racial categories.

\begin{longtable}[]{@{}lr@{}}

\caption{\label{tbl-race-summary}Distribution of patients by race in the
analytic cohort.}

\tabularnewline

\toprule\noalign{}
race & Count \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
American Indian/Alaska Native & 38 \\
Asian or Pacific Islander & 506 \\
Black & 1028 \\
Unknown & 12 \\
White & 4124 \\

\end{longtable}

\subsection{Exploratory Summary: Year of
Diagnosis}\label{exploratory-summary-year-of-diagnosis}

We also summarized the distribution of patients by year of cancer
diagnosis to understand the temporal spread of the study cohort. This
information helps contextualize potential changes in treatment patterns,
diagnostic advancements, or survival outcomes over time.
Table~\ref{tbl-year-summary}. Frequency distribution of patients by year
of diagnosis.

\begin{longtable}[]{@{}lr@{}}

\caption{\label{tbl-year-summary}Distribution of patients by year of
diagnosis.}

\tabularnewline

\toprule\noalign{}
diagnosis\_year & Count \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2010 & 821 \\
2011 & 889 \\
2012 & 960 \\
2013 & 972 \\
2014 & 962 \\
2015 & 1104 \\

\end{longtable}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Bar plot displaying the distribution of patients by year of diagnosis.
  The figure shows a relatively even spread across the years, with some
  fluctuations that may reflect data entry trends or cohort recruitment
  variability.
\end{enumerate}

\begin{figure}

\centering{

\includegraphics[width=7.29in,height=\textheight]{../../../results/edafigures/edayear.png}

}

\caption{\label{fig-year-diagnosis}Bar plot showing the distribution of
patients by year of diagnosis.}

\end{figure}%

\subsection{Exploratory Summary: Cause-Specific Death
Classification}\label{exploratory-summary-cause-specific-death-classification}

We summarized the distribution of outcomes according to the SEER
cause-specific death classification. The majority of patients in the
dataset (77.5\%) were classified as having died due to the primary
cancer diagnosis, while 22.5\% were either alive or had died from other
causes. This outcome classification was used as the primary event
definition in survival models throughout the analysis.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Distribution of patients by cause-specific death classification.
\end{enumerate}

\begin{longtable}[]{@{}lrr@{}}

\caption{\label{tbl-cause-death}Cause-specific death classification
among patients in the cohort.}

\tabularnewline

\toprule\noalign{}
cause\_specific\_death & Count & Percentage \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Alive or dead of other cause & 1287 & 22.5 \\
Dead (attributable to this cancer dx) & 4421 & 77.5 \\

\end{longtable}

\subsection{Univariate Analysis}\label{univariate-analysis}

To evaluate whether survival outcomes varied over time, we conducted a
bivariate Cox proportional hazards analysis using year of diagnosis as
the predictor. The year 2010 was used as the reference category. Hazard
ratios (HRs) for each subsequent year (2011 to 2015) were calculated to
assess whether patients diagnosed in later years had different risks of
death compared to those diagnosed in 2010. As shown in
Table~\ref{tbl-year-diagnosis}, all HRs were close to 1, and none of the
comparisons reached statistical significance (all p \textgreater{}
0.05), indicating that there was no significant year-to-year variation
in survival over the study period. These findings suggest a relative
consistency in survival outcomes for patients diagnosed between 2010 and
2015 within this dataset.

\begin{longtable}[]{@{}lrrrr@{}}

\caption{\label{tbl-year-diagnosis}Hazard ratios from Cox regression
evaluating year of diagnosis and survival.}

\tabularnewline

\toprule\noalign{}
term & estimate & std.error & statistic & p.value \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
diagnosis\_year2011 vs.~2010 & 0.981 & 0.054 & -0.368 & 0.713 \\
diagnosis\_year2012 vs.~2010 & 0.984 & 0.052 & -0.316 & 0.752 \\
diagnosis\_year2013 vs.~2010 & 1.000 & 0.053 & -0.001 & 1.000 \\
diagnosis\_year2014 vs.~2010 & 0.978 & 0.053 & -0.429 & 0.668 \\
diagnosis\_year2015 vs.~2010 & 0.982 & 0.052 & -0.349 & 0.727 \\

\end{longtable}

\subsection{Random Forest Model: Predicting Cause-Specific
Death}\label{random-forest-model-predicting-cause-specific-death}

We implemented a random forest classifier to predict whether a patient's
death was attributable to cancer using SEER registry variables. The
dataset was preprocessed with downsampling to address class imbalance.
We used 5-fold cross-validation to tune hyperparameters and selected the
best model based on ROC AUC. The model was trained on 80\% of the data
and evaluated on a held-out 20\% test set.

The top predictors, based on variable importance, included lung and
liver metastases, age group, and year of diagnosis.

Figure~\ref{fig-vip-rf} .Variable importance plot from the final tuned
random forest model. The top 10 predictors are displayed, with
mets\_lung and mets\_liver being the most influential variables for
predicting cause-specific death.

\begin{figure}

\centering{

\includegraphics[width=8in,height=\textheight]{../../../results/figures/vip_plot_rf_balanced.png}

}

\caption{\label{fig-vip-rf}Top 10 most important variables from the
random forest model predicting cause-specific death.}

\end{figure}%

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Confusion matrix summarizing the random forest model's classification
  performance on the test set. The model shows better performance in
  identifying cancer-attributable deaths than non-cancer deaths,
  although overall accuracy was limited.
\end{enumerate}

\begin{table}

\caption{\label{tbl-rf-cm}Confusion matrix from the random forest model
on the test set.}

\centering{

\begin{verbatim}
$table
          Truth
Prediction  No Yes
       No  152 385
       Yes  92 447

attr(,"class")
[1] "conf_mat"
\end{verbatim}

}

\end{table}%

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Area under the receiver operating characteristic curve (ROC AUC) from
  the random forest model. The model achieved an AUC of approximately
  0.39, indicating limited discriminative performance, possibly due to
  overlapping features and class imbalance.
\end{enumerate}

\begin{table}

\caption{\label{tbl-rf-auc}ROC AUC score of the final random forest
model.}

\centering{

\begin{verbatim}
# A tibble: 1 x 3
  .metric .estimator .estimate
  <chr>   <chr>          <dbl>
1 roc_auc binary         0.394
\end{verbatim}

}

\end{table}%

\subsection{LASSO Cox Regression for Variable
Selection.}\label{lasso-cox-regression-for-variable-selection.}

To identify the most influential predictors of cancer-specific survival,
we applied LASSO (Least Absolute Shrinkage and Selection Operator)
penalized Cox regression using the glmnet package. This approach helps
in variable selection by shrinking less informative coefficients to zero
while retaining strong predictors. We performed 10-fold cross-validation
to select the optimal penalty parameter (λ), minimizing the partial
likelihood deviance.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Selected variables from the LASSO model using the optimal penalty
  parameter (λ.min). These are the predictors that remained in the final
  model with non-zero coefficients.
\end{enumerate}

\begin{longtable}[]{@{}lr@{}}

\caption{\label{tbl-lasso-vars}Non-zero coefficients from the LASSO Cox
regression model.}

\tabularnewline

\toprule\noalign{}
variable & coefficient \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
mets\_lungYes & 0.3323 \\
mets\_liverYes & 0.3862 \\
mets\_boneYes & 0.2833 \\
mets\_brainYes & 0.5570 \\
age\_group25-29 years & -0.1979 \\
age\_group30-34 years & -0.1054 \\
age\_group35-39 years & -0.2170 \\
age\_group40-44 years & -0.3530 \\
age\_group45-49 years & -0.1630 \\
age\_group50-54 years & -0.0678 \\
age\_group55-59 years & -0.0226 \\
age\_group60-64 years & 0.0159 \\
age\_group70-74 years & 0.2417 \\
age\_group75-79 years & 0.3073 \\
age\_group80-84 years & 0.4241 \\
age\_group85+ years & 0.6496 \\
raceBlack & 0.3345 \\
raceAsian or Pacific Islander & -0.0173 \\
raceAmerican Indian/Alaska Native & 0.1343 \\
median\_income\$40,000 - \$44,999 & 0.0456 \\
median\_income\$50,000 - \$54,999 & 0.0571 \\
median\_income\$65,000 - \$69,999 & 0.0468 \\
median\_income\$70,000 - \$74,999 & 0.0038 \\
diagnosis\_year2013 & 0.0283 \\
diagnosis\_year2014 & -0.0097 \\

\end{longtable}

\newpage{}

\end{document}

### 2025-04-18 products/manuscript/supplement/Supplementary-Material.qmd
Part 5 V3

@@
 {{< pagebreak >}}
-
-# Additional Method Details
-
-Often, the main manuscript only allows for an overview description of the methods.
-Use the supplement to describe all your methods, models and approaches in a lot of detail.
-Reference specific parts of your code as needed.
-
-{{< pagebreak >}}
-
-# Additional results
-
-Show additional results here. Those can be some useful exploratory/descriptive figures or tables, or results from additional analyses that didn't make it into the main text.
 

### 2025-04-26 reviews/Joseph.project-review-template.qmd
Par 5 review. Hope & Joseph

---
title: The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort  
author: Asmith Joseph
date: 4/25/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project:The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort

Name of project author(s): Mohammed Zuber

Name of project reviewer: Asmith Joseph

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/renders into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The background section provides a clear foundation: incidence, stages, and survival. 
Some suggestions to make it more concise: add one line on why site-specific metastasis data are still lacking, use a single citation style, combine duplicate stats, trim the symptom list, and cite a few prior SEER studies to show how you advance the field.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments
Your core question and its connection to SEER data are straightforward and well-defined. 
2 suggestions: 
* Split the aim into two succinct questions, one on clinicopathologic features and one on survival.
* Spell out the exact survival metrics you’ll analyze (e.g., overall vs. cancer-specific survival, median vs. five-year rates).

### Summary assessment
* question/hypotheses unclear
* question/hypotheses somewhat explained
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments
The data description is comprehensive: you clearly outline where records were extracted and specify precise inclusion criteria.
Both demographic and clinical variables are defined and categorized in detail.
While the SEER source is cited correctly, no standalone codebook or tailored data dictionary is provided. 
To ensure complete transparency and reproducibility, include or reference the SEER data dictionary or supply a project-specific codebook that maps every variable and category.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The data-cleaning workflow is rigorous and fully reproducible. You define inclusion and exclusion criteria precisely, standardize variable types consistently, and share code and raw data in organized folders. Concise summary tables and informative visualizations support your exploratory analysis well.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments
Applying Kaplan–Meier curves, Cox proportional-hazards models, and a Random Survival Forest (RSF) with 500 trees, proper factor conversion, and impurity-based variable importance is well suited to right-censored survival data: the two-stage Cox approach cleanly separates unadjusted from adjusted effects, and the RSF captures nonlinearities and interactions.

Some suggestions: 
* report checks of the Cox models’ proportional-hazards assumption and explain how missing covariate data were handled
* justify your predictor selection criteria and provide internal validation for the RSF (e.g., cross-validation or out-of-bag error estimates)
* given the censoring structure, either justify using a confusion matrix or adopt a more appropriate performance metric.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments
All presentation criteria are satisfied:
* Abbreviations and symbols are defined in legends or footnotes
* Font styles, sizes, and line weights are consistent, legible at journal-print Scale, and distinguishable in greyscale for color-blind readers
* Captions succinctly convey key findings
* Table columns are aligned with appropriate numerical precision and footnoted for significance
* Every figure and table is cited in a logical order with matching numbering

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments
* The discussion clearly interprets your key findings, showing how lung, brain, and liver metastases uniquely affect survival and situating these results within the broader literature on aggressive visceral spread. 
* You appropriately highlight the study’s strengths and openly address its limitations. 
*Your conclusions follow logically from the data and propose valuable next steps, such as integrating molecular and longitudinal treatment information.

Some Suggestions:
*Briefly discuss how unmeasured clinical factors might bias your hazard estimates and outline specific ways these insights could inform clinical decision-making.

### Summary assessment
* strong, complete and clear discussion

## Further comments

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

Some Final Suggestions: 
* Consider adding a brief sensitivity analysis, such as comparing complete-case versus imputed datasets, to confirm the robustness of your findings. 
* Also, include a statement in the Discussion on how metastatic-site risk stratification could inform treatment guidelines or multidisciplinary care pathways.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Upon reviewing the project files, I noticed that some folders in the main "Asset" directory do not seem relevant to the project. Remove these to avoid confusion and improve organization. 
On the positive side, the other folders, such as code, products, results, and data, are well-organized and structured. 
This setup allows for easy navigation and makes it straightforward to reproduce the analysis. 
Overall, the project is structured to support efficient workflow and ensure that all necessary files are accessible.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

The project is very well documented. I can clearly understand each step of the analysis, the reasoning behind each decision, and the logic behind each line of code.
The code is appropriately commented on, and the Rmd files provide enough detail to follow the workflow without confusion.
The thorough documentation ensures that anyone reviewing the project can easily comprehend the process and replicate the analysis.

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Yes, all results are fully reproducible. The documentation clearly explains how to reproduce the analysis, and everything works seamlessly without the need for manual intervention. 
I was able to re-run the entire analysis without making any edits or adjustments. 
The workflow is well-organized, making it easy to replicate the results and follow the steps as outlined.

### Summary assessment
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments
* The study thoroughly explores clinicopathological features and survival outcomes in stage IV endometrial cancer, focusing on metastatic sites (lung, liver, bone, and brain).
* It combines Kaplan-Meier and Cox models with a Random Survival Forest (RSF) model, offering robust survival estimates and insights into predictor importance.
* The research questions are clearly defined and fully addressed. 
* The hypothesis linking metastasis to worse survival is explored in depth. Factors like age, race, and income are considered, and survival disparities across these groups are also examined.

* Overall, the study is well-structured and comprehensive and effectively uses both statistical and machine-learning methods to answer the research questions.

### Summary assessment
* strong level of thorougness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._

The project is well-executed and thoroughly addresses the key aspects of survival analysis in stage IV endometrial cancer. 
* Excellent work. It's clear that you put a lot of effort into this research. Thank you for sharing it with us.

### 2025-04-26 reviews/zuber-review-byhopeg.qmd
Par 5 review. Hope & Joseph

---
title: Project Review Template 
author: Hope Grismer
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: The Clinicopathological Features and Survival Outcomes of Patients with Different Metastatic Sites in Stage IV Endometrial Cancer: A SEER-Based Cohort Study

Name of project author(s): Mohammed Zuber

Name of project reviewer: Hope Grismer

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/renders into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The manuscript provides strong background information, including clear statistics about incidence and survival rates of endometrial cancer. Prior studies are referenced appropriately. The motivation for the study — understanding survival outcomes by metastasis site — is well established and timely. The importance of investigating metastatic patterns is clearly justified given the survival disparities highlighted.

Minor note: A little more explicit tie-in about what gaps remain in the current literature would make the motivation even stronger (e.g., “Previous studies have not systematically compared brain, liver, bone, and lung outcomes in a population-based cohort”).

### Summary assessment (PICK ONE, DELETE THE OTHERS)

-   strong contextualization and motivation

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The research question is clearly stated and fully aligned with the study goals. Both the question ("What are the survival outcomes across metastatic sites?") and the hypothesis ("different sites have different survival implications") are logically connected to the data being analyzed. Great job stating not only the expected differences but how they might arise (tumor behavior, treatment responses).

### Summary assessment

-   question/hypotheses fully clear

## Data description

How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The SEER database isv**well described**, including when it started capturing metastasis-specific data (2010), and the exact fields pulled for analysis.

Minor nitpick: There’s a typo where "brain metastasis" is listed twice in the extracted variables section.

### Summary assessment

-   source and overall structure of data well explained

## Data wrangling and exploratory analysis

How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The cleaning steps are reasonably explained:

-   Exclusions (age \<18, non-endometrial deaths, early-stage cancers) are sensible

-   Variable formatting (categorical to factors) is described

-   Missingness handling is addressed

The exploratory analysis includes descriptive stats, tables, and good plots for survival time, metastasis site, and demographics.

Minor room for improvement:

-   Could use slightly more commenting in the code (especially in processingcodemz)

-   Exploratory analysis focuses mainly on survival; a few more EDA plots (e.g., cross-tabulation of metastasis by race) could deepen understanding.

### Summary assessment

-   some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The statistical analysis is appropriate and well implemented:

-   Kaplan-Meier curves are correctly used for unadjusted survival comparisons

-   Bivariate and multivariable Cox models are used correctly

-   Random Survival Forest model is a strong (and justified) complement to traditional modeling

Model performance (AUC = 0.74) is respectable. Liked your choice to focus on cause-specific survival rather than all-cause mortality.

### Summary assessment

-   strong and reasonable analysis

## Presentation

How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Tables and figures are clear and logically described in the text, even though the actual figures were not attached to the Word document.\
(From reading the code, I can tell they would render well — e.g., ggplot2 survival curves, Random Forest variable importance plots.)

A bit more consistent labeling of axes/titles would elevate figures to true publication quality.

### Summary assessment

-   results are very well presented (tiny room for improvement)

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion is thorough and well contextualized:

-   Main findings (brain/liver worse prognosis) are clearly summarized

-   Racial and age disparities are thoughtfully discussed

-   Strengths (population-based sample) and limitations (no treatment data) are acknowledged

The interpretation is responsible — no overstatement of the findings. References are integrated nicely.

### Summary assessment

-   strong, complete and clear discussion

## Further comments

*Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project.*

# Overall project content evaluation

Evaluate overall features of the project by filling in the sections below.

## Structure

Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

-   Three main Quarto files corresponding to cleaning, EDA, and analysis

-   No junk files

-   Clear naming conventions

-   Could organize some things into subfiles and folders???

### Summary assessment

-   Well structured

## Documentation

How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

The manuscript is extremely well documented.\
The code is moderately documented

### Summary assessment

-   decently documented with some gaps

## Reproducibility

Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

-   All code files are provided

-   Data source (SEER) is identified

-   GitHub repo link is included

### Summary assessment

-   fully reproducible without issues

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

-   Both traditional statistics and machine learning were applied

-   Multiple predictors were considered

-   Sensitivity to race, age, and income differences were discussed

-   Strengths and limitations were there too

### Summary assessment

-   strong level of thorougness

## Further comments

Thank you for an amazing project! I really was inspired in getting to look yours over. It was super well-rounded project– very close to publication-level. I especially liked the depth and strength of writing in your manuscript.

### 2025-05-01 products/manuscript/Manuscript.qmd
Part 6

@@
 However, this data specifically pertains to stage IV lung cancer patients.[@Mao2020] Patients with advanced endometrial cancer face poor survival outcomes, with metastatic disease being the leading cause of mortality.
-[@Mao2020] Therefore, this study aimed to utilize data from the Surveillance, Epidemiology, and End Results (SEER) database to analyze survival outcomes in patients diagnosed with stage IV endometrial cancer.
+[@Mao2020].
+
+While prior studies have reported outcomes by site-specific metastases, few have systematically compared survival outcomes across brain, liver, bone, and lung metastases in a large population-based cohort of stage IV endometrial cancer patients.
+This gap limits a comprehensive understanding of prognosis across different metastatic patterns.
+
+Therefore, this study aimed to utilize data from the Surveillance, Epidemiology, and End Results (SEER) database to analyze survival outcomes in patients diagnosed with stage IV endometrial cancer.
 The primary objective was to determine whether the presence of lung/brain/bone/liver metastases in adults with stage IV endometrial cancer is associated with worse survival outcomes compared to patients without any metastasis.
